Neural spiking signatures predict behavioral phenotypes of cerebellar movement disorders

Version 1

Abstract

Abstract

The cerebellum contributes to a diverse array of motor conditions including ataxia, dystonia, and tremor. The neural substrates that encode this diversity are unclear. Here, we tested whether the neural spike activity of cerebellar output neurons predicts the phenotypic presentation of cerebellar pathophysiology. Using in vivo awake recordings as input data, we trained a supervised classifier model to differentiate the spike parameters between mouse models for ataxia, dystonia, and tremor. The classifier model correctly predicted mouse phenotypes based on single neuron signatures. Spike signatures were shared across etiologically distinct but phenotypically similar disease models. Mimicking these pathophysiological spike signatures with optogenetics induced the predicted motor impairments in otherwise healthy mice. These data show that distinct spike signatures promote the behavioral presentation of cerebellar diseases.

Introduction

There exists a historical and rich curiosity in understanding the role of the cerebellum in movement, dating back to the pioneering work of Flourens (1), with an equally long interest in investigating how it alters movement (2, 3). From these earlier studies it was clear that a defective cerebellum causes a range of devastating problems in the ability to control voluntary, intentional actions including coordination, posture, and balance. Accordingly, in humans, the equivalent defects had clear pathophysiological consequences. Patients with various cerebellar lesions showed loss of precise motor coordination, which, as the consequence of disease, is referred to as ataxia, whereas some subjects displayed overt and sometimes exaggerated oscillatory movements, consistent with tremor (2–4). In other cases, the examiner would report uncontrolled muscle contractions, which is now a contributing feature of dystonia (5). These classic observations underscored the heterogeneity of motor disturbances caused by cerebellar dysfunction. But, even then, the question of how the cerebellum creates such behavioral diversity was already imminent. Despite major advances in understanding the basic anatomy and circuitry of the cerebellum, it is still unclear how disease behaviors emerge of these circuits. In this regard, the specific substrates that underlie each disorder could hold the key for improving the design of therapies and treatments.

Consistent with the outcomes of removing or lesioning the cerebellum in pigeons, dogs, monkeys, and humans (2–4), the heterogeneity of cerebellar movement disturbances was later confirmed in genetic mouse models. Much of the current excitement about the cerebellum spawned from these initial genetic models because of the overt motor disturbances that were caused by spontaneous mutations in genes that are now known to be associated with cerebellar development or cerebellar degeneration (6, 7). Prior to genetic sequencing though, mutant mice such as hot-foot (8), weaver (9), trembler (6), waddles (10), staggerer (11), stumbler (12), tottering (13), and lurcher (14) were named according to phenotype of their abnormal movements, which are as diverse as their names imply. Modern techniques and approaches now aim at determining the mechanisms and roles of the cerebellum as they relate to driving different behaviors in these classic models (15, 16). Together, data generated from these different mouse models have cultivated an interest in identifying whether discrete functional features in the cerebellar circuit are the root cause of disease-related behaviors. In this regard, more than half a century since the first descriptions of cerebellar mutants, a core question remains unanswered but hotly debated: how does cerebellar circuit dysfunction lead to unique motor disturbances?

To begin addressing how cerebellar circuits generate behavioral diversity in disease, we used an intersectional genetics approach to mark, map, and manipulate specific types of synapses in the cerebellum. Our approach silences genetically defined populations of synapses by selectively deleting the genes that encode the vesicular GABA transporter, Vgat, or the type 2 vesicular glutamate transporter, Vglut2, from genetically targeted populations of cells. Loss of GABAergic neurotransmission from Purkinje cells, which provide the sole output of the cerebellar cortex, caused uncoordinated movements and disequilibrium that were indicative of ataxia (Figure 1A) (17). In contrast, eliminating glutamatergic neurotransmission from climbing fibers, which synapse onto Purkinje cells, caused twisting postures and hyperextended limbs that is consistent with dystonia (Figure 1A) (18). In addition, systemic injection of the beta-carboline alkaloid drug harmaline caused hyperactivation of climbing fibers and rhythmic bursting spike activity in Purkinje cells as well as a severe tremor. Optogenetically modulating Purkinje cell to cerebellar nuclei projections in a bursting pattern induced oscillatory tremor movements (Figure 1A) (19, 20). Together, these studies established that different manipulations of Purkinje cell inputs or outputs, and consequently Purkinje cell and nuclei neuron spike signals, can cause diverse behavioral deficits as they relate to disease (18, 20). Given the heterogeneity and even comorbidity of these behaviors in a single disease model (21), the main question that arises is, do these cerebellar neural signals represent unique pathophysiological signatures that can drive the predominant behavioral defects used to characterize different motor diseases? Here, we aim to resolve this question to provide insight into the origin of cerebellar movement disorder presentation.

Supervised classifier model predicts mouse phenotype based on spike signatures.

A. Schematic of network changes causing motor impairments in mouse models for ataxia, dystonia, and tremor. Dotted lines indicate lack of neurotransmission. Color-coded lines indicate primary affected cell type. B. Example of spike firing rate averaged over previous 50 ms at each occurring spike for the 5 s spike train in B’. B’’ 1 s spike train for the duration indicated in the square box in B’. C. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train used in the classification model. D. Classifier model based on training data set (control: n = 29 cells, N = 9 mice; ataxia: n = 19, N = 5; dystonia: n = 19, N = 8; tremor: n = 21, N = 6). E. Predicted phenotype of mice based on spike properties in validation data set (control: n = 30 cells, N = 9 mice; ataxia: n = 18, N = 5; dystonia: n = 17, N = 9; tremor: n = 18, N = 5). Categories on X-axis indicate the origin of the recorded neurons. F. and G. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons.

PC = Purkinje cell; CN = cerebellar nuclei; IO/CF = inferior olive/climbing fiber.

Results

Spike signatures are different between archetypal ataxia, dystonia, and tremor models

We first compared the spike train activity between mouse models for ataxia (L7Cre;Vgatfl/fl, (17, 20)), dystonia (Ptf1aCre/+;Vglut2fl/fl, (18, 22)), and tremor (harmaline injection, (20, 23)) (Figure 1A). We identified these three mouse models as archetypal representations of their respective cerebellar motor disease phenotypes because their overt behavioral motor abnormalities are caused by cerebellum-specific manipulations that do not cause changes in the gross anatomy, cell morphology, or cell survival of the adult cerebellum. These models also exhibit remarkably reliable, severe, and penetrant ataxic, dystonic, and tremor symptoms within each respective group. Video 1 provides examples of the varied and overt motor abnormalities of the mouse models used. Supplemental table 1 summarizes the reported behavioral impairments from prior characterization of these archetypal mouse models.

We hypothesized that these differences in behavioral abnormalities were accompanied by specific changes in the spike train patterns in the cerebellum. We performed in vivo electrophysiology recordings of the spike activity in cerebellar nuclei neurons of the interposed nucleus (Figure 1B,C), in awake, head-fixed mice with overt motor phenotypes and control animals. We focused our recordings on the interposed nucleus based on the hypothesis that the cerebellum communicates ongoing motor signals to other brain regions through this nucleus (24) and deep brain stimulation (DBS) of this region successfully reduces motor impairments in mouse models of ataxia (25), dystonia (18), and tremor (20). We described the spike train firing features using twelve parameters that summarized the spike train rate (Supp. Figure 1A-C), irregularity (Supp. Figure 1D-F), pauses (Supp. Figure 1G-I), and rhythmicity (Supp. Figure 1J-L). In agreement with our hypothesis, we found a significant difference between at least two archetypal groups for each of the twelve spike train parameters. However, none of the twelve parameters showed a statistical difference between all four groups (control, ataxia, dystonia, and tremor), suggesting that the difference between spike train signatures relied on a combination of multiple spike train parameters.

We therefore set out to find a set of parameters and specific threshold values that best differentiated the spike train parameters between the archetypal groups. We trained a supervised classification learning model based on the spike train parameters of each mouse model (Supp. Figure 1A-L). The classifier model identified the coefficient of variance (CV, Supp. Figure 1D) of interspike intervals (ISI) as the best differentiator of neurons recorded in ataxic and control mice (lower CV) from neurons recorded in dystonic and tremoring mice (higher CV). CV is a measure of variability that increases due to pauses between spikes or fluctuations in the firing rate (Figure 1B) and is unusually elevated in our dystonic and tremor mouse models (Figure 1B”). Next, the classifier model identified CV2 (26) (Supp. Figure 1E) as the best differentiator of neurons recorded in ataxic mice (lower CV2) from control mice (higher CV2). CV2 measures the irregularity of interspike intervals with less influence from the overall firing rate. It is extraordinarily low in the highly regular spike activity observed in our ataxia model (Figure 1B’’). Last, the classifier model identified that the median ISI-1 (Supp. Figure 1C), which measures the median instantaneous firing rate, as the best differentiator of neurons from the dystonic mice (lower median ISI-1) compared to those recorded in mice with tremor (higher median ISI-1) (Figure 1C). These findings suggest that changes in a combination of three parameters (CV, CV2, and median ISI-1) may explain the difference in the behavioral presentation of mouse models of cerebellar disease.

We then validated whether this classifier model could reliably predict the behavioral phenotypes of the archetypal mouse models based solely on the spiking activity of single cerebellar nuclei neurons. We tested this by predicting the behavioral phenotype based on CV, CV2, and median ISI-1 values calculated based on a different set of recordings. We found that the classifier model indeed predicted the different predominant phenotypes of each archetypal mouse model correctly for the majority of neurons in the validation data set (control: 80% of neurons recorded in a control mouse had a control signature; ataxia: 78%; dystonia: 65%; tremor: 56%) (Figure 1E). Our analyses also showed that most neurons of a specific signature were recorded from the mouse model with that phenotype (control: 59% of neurons with a control signature were recorded in a control mouse; ataxia: 89%; dystonia: 65%; tremor: 73%) (Figure 1E-G). These data support the hypothesis that spike signatures are reliably different between archetypal mouse models of distinct motor behaviors that mimic human disease symptoms.

Multiple spike train signatures can lead to ataxic motor impairments

Next, we set out to identify whether the spike train signatures for the archetypal mouse models are shared across mouse models with different manipulations but similar phenotypes. We started by classifying the spike train parameters in a mouse model of ataxia caused by a poly-glutamate expansion in the Atxn1 gene, Atxn1154Q/+ mice, that causes spinocerebellar ataxia type 1 (SCA1). We measured the spike train patterns early in the disease progression (27) (Figure 2A-B). Despite the presence of an ataxic phenotype at this age (Supp. Table 1), we observed that most Atxn1154Q/+neurons exhibited a control signature (14/24) and that the second most numerous spike train signature was a dystonia signature (8/24).

Spike signatures in ataxia models with different etiologies.

A 5 s example spike trains (and 1 s inset) of representative cerebellar nuclei neurons recorded in each mutant mouse model. Left: example cell from early disease progression (4-month-old) ataxic Atxn1154Q/+ mouse. Right: example cell from late disease progression (11-month-old) ataxic L7Cre;Ank1fl/fl mouse. B. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train of the example cells in A. C. Proportion of predicted spike signatures in each of the mouse models. D. and E. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons. Control: N = 11, n= 55; Atxn1154Q/+: N = 4, n = 24; L7Cre;Ank1fl/fl: N = 3, n = 14. C-E. are based on classifier model in Figure 1D.

We reasoned that the high proportion of cells exhibiting a control signature in Atxn1154Q/+ mice might be because the disease-causing genetic abnormality affects the whole body and, consequently, the ataxic symptoms may result from dysfunction in multiple nodes in the motor circuit. We therefore also investigated a mouse model with a cerebellum-specific loss of the gene Ank1 (10-12 month old L7Cre;Ank1fl/flmice, Figure 2A), which causes an adult-onset degenerative ataxia (28). In this mouse model, we found that more than half of the neurons exhibited an abnormal spike train signature, with the majority classified with the dystonia spike train signature (7/14).

Our classifier model differentiates the control and dystonia signatures from each other based on the CV parameter. Previous studies have suggested that the difference between ataxia and dystonia symptoms may be caused by the relative difference in CV value (15, 29, 30). Indeed, the distribution of CV values appeared to be lower in the ataxic Atxn1154Q/+ (0.59 ± 0.04 (mean ± SEM)) and L7Cre;Ank1fl/fl(0.61 ± 0.07) mouse models (Figure 2D-E) than in our archetypal dystonia model (1.15 ± 0.11) (Ptf1aCre/+;Vglut2fl/fl, Figure 1F-G). We used a one-way ANOVA followed by a Tukey post-hoc analyses and found that the difference in CV between the ataxia models and the dystonia models was statistically significant (Atxn1154Q/+vs. Ptf1aCre/+;Vglut2fl/fl: p<0.0001; L7Cre;Ank1fl/flvs. Ptf1aCre/+;Vglut2fl/fl: p<0.0001), but we did not observe a difference in CV between the two ataxia models (Atxn1154Q/+vs L7Cre;Ank1fl/fl: p=0.9739). Together, these findings suggest that there are multiple cerebellar spike train signatures that can lead to ataxic symptoms that both differ from the dystonia signature in their relative level of regularity; unusually regular spike train patterns are seen in L7Cre;Vgatfl/fl mice (Figure 1) and intermediate irregular spike train patterns are observed in Atxn1154Q/+ and L7Cre;Ank1fl/flmouse models (Figure 2).

Dystonic spike train signatures are shared across etiologically distinct dystonia models

Next, we set out to identify whether the dystonia signature was shared across mice with overt dystonic phenotypes. We included two dystonia models in our analyses; Pdx1Cre;Vglut2fl/fl mice exhibit mild dystonic symptoms only in response to a stressful event (31, 32) (Figure 3A-B), whereas mice that receive ouabain applied to the surface of the cerebellum exhibit continuous severe dystonic features (33, 34) (Figure 3A-B). We found that the proportion of neurons that exhibited the control signature was greater in Pdx1Cre;Vglut2fl/flmice than in ouabain infusion mice (11/21 vs. 7/20) (Figure 3C). Moreover, the proportion of neurons with a dystonia signature was lower in Pdx1Cre;Vglut2fl/flmice than in ouabain infusion mice (9/21 vs. 11/20) (Figure 3C). Therefore, the mice with a more severe dystonia phenotype had a smaller proportion of cells matching the control signature and a greater proportion of cells matching the dystonia signature. In addition to the proportional differences in cells with a control and dystonia signature, we found that the distribution of CV values differed between the Pdx1Cre;Vglut2fl/fl (0.61 ± 0.07) and ouabain infusion mice (1.00 ± 0.15) (Figure 3D-E). We quantified this difference by comparing these two mouse models to each other and the archetypal dystonic model (Ptf1aCre/+;Vglut2fl/fl) using a one-way ANOVA followed by a Tukey post-hoc analysis. We found that the CV was lower in Pdx1Cre;Vglut2fl/fl mice with mild dystonic features compared to the ouabain infusion mice (p=0.0482) and Ptf1aCre/+;Vglut2fl/flmice (p=0.0037) with severe dystonic features. There was no significant difference between the two models with severe dystonic features (p=0.5940). These findings show that the dystonia signature is shared across etiologically distinct dystonia models and that the relative severity of symptoms may be explained by the proportion of neurons with a dystonia and control signature (Figure 3).

Spike signatures in dystonia models with different etiologies.

A 5 s example spike trains (and 1 s inset) of representative cerebellar nuclei neurons recorded in each mutant mouse model. Left: example cell from mildly dystonic Pdx1Cre;Vglut2fl/flmouse. Right: example cell from severely dystonic cerebellum ouabain infusion mouse. B. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train of the example cells in A. C. Proportion of predicted spike signatures in each of the mouse models. D. and E. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons. Control: N = 6, n= 28; Pdx1Cre;Vglut2fl/fl: N = 4, n = 21; cerebellum ouabain infusion: N = 4, n = 20. C-E. are based on classifier model in Figure 1D.

Tremor spike train signatures are shared across etiologically distinct tremor models

Next, we set out to identify whether the tremor signature is shared across mice that exhibit severe oscillatory tremors. To test this, we used Car8wdl/wdl mice. We previously showed that Car8wdl/wdlmice exhibit a severe oscillatory tremor that can be reduced to control levels by treatment with the beta blocker drug propranolol (35), which is used to treat tremor in human patients. Car8wdl/wdl mice also exhibit ataxia and dystonia, which are not normalized by propranolol (10, 21, 25, 35). We found that, before treatment, Car8wdl/wdlmice exhibited a large proportion of neurons with the tremor signature (12/19). However, in propranolol-treated Car8wdl/wdl mice the proportion of neurons with the tremor signature was diminished while control (6/15) and dystonia (6/15) signatures were the most common signatures (Figure 4C). The classifier model differentiates between the dystonia and tremor signature based on the median ISI-1 value and, indeed, we find a shift towards lower median ISI-1 values in propranolol-treated Car8wdl/wdl mice compared to untreated Car8wdl/wdl mice (Figure 4D-E). Together, these findings suggest that the tremor signature is shared across etiologically distinct mouse models of tremor and that the tremor-reducing drug propranolol also reduces specific spike train features contributing to the tremor signature (Figure 4).

Spike signatures in tremor models with different etiologies.

A 5 s example spike trains (and 1 s inset) of representative cerebellar nuclei neurons recorded in each mutant mouse model. Left: example cell from Car8wdl/wdl mouse with complex phenotype including severe tremor. Right: example cell from Car8wdl/wdl mouse with complex phenotype treated with propranolol to treat tremor. B. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train of the example cells in A. C. Proportion of predicted spike signatures in each of the mouse models. D. and E. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons. Control: N = 4, n= 14; Car8wdl/wdl: N = 9, n = 19; Car8wdl/wdl + propranolol: N = 6, n = 15. C-E. are based on classifier model in Figure 1D.

Different spike signatures can be generated by the same canonical cerebellar circuit

Next, we investigated whether the different spike signatures could be generated by the same canonical cerebellar circuit. We reasoned that the observed spike train signatures were not dependent on developmental changes in circuit connectivity (L7Cre;Vgatfl/fl; Ptf1aCre/+;Vglut2fl/fl; Pdx1Cre;Vglut2fl/fl), neurodegeneration (Atxn1154Q/+; L7Cre;Ank1fl/fl), the effects of transgene expression (Atxn1154Q/+) or frameshifting (Car8wdl/wdl), or drug effects on the nervous system outside of the cerebellum (harmaline, ouabain, propranolol). We expressed a light-sensitive cation channel opsin, channelrhodopsin (ChR2), in Purkinje cells (L7Cre;ROSA26loxP-STOP-loxP-EYFP-ChR2 mice, hereon referred to as L7Cre;R26ChR2 mice). Upon light activation, Purkinje cell firing rate increases (Supplemental Figure 2A-B) and nuclei neurons are inhibited via the GABAergic neurotransmission from Purkinje cells (Figure 5A-C). We leveraged this circuit connectivity and optogenetic strategy to induce distinct spike signatures in cerebellar neurons. We chose three stimulation paradigms. The ataxia stimulation was a continuous 50 Hz square pulse stimulation (mimicking the lack of modulation by Purkinje cells in our degenerative and genetically silenced ataxia models); the dystonia stimulation was a 50 Hz square pulse stimulation that was on for at least 500 ms with a 75% chance of a 250 ms pause between periods of stimulation (to induce the slow and irregular spike train features of our defined dystonia signature); and the tremor stimulation was a 10 Hz sinusoidal stimulation (to induce the fast, irregular, and rhythmic spike train features of our defined tremor signature). We performed in vivo recordings in awake, head-fixed mice, and confirmed that each of the optogenetic stimulation paradigms caused distinct changes in Purkinje cell spiking activity (Supplemental Figure 2). The different stimulation paradigms also induced distinct changes in the spike train features of cerebellar nuclei neurons (Figure 5D). We then compared the resultant spike train features to the defined thresholds of our classifier model. We found that the ataxia stimulation paradigm induced cells to behave within the parameters of a variety of our defined spike train signatures, including ataxia (2/9), dystonia (2/9), and tremor (2/9). This is in line with our findings that the ataxia signature is more variable across ataxia mouse models (Figure 2). Importantly, we found that fewer than half of the cells recorded with the ataxia stimulation exhibited a control signature (Figure 5F), similar to the proportional distribution of cells in other mouse models with severe motor phenotypes (Figure 2C, 3C, 4C). Additionally, we found that the dystonia and tremor stimulation reliably induced spike signatures that our classifier model defined as dystonic (7/7 cells) or as tremor signatures (9/9 cells), respectively (Figure 5F). These data indicate that no anatomical changes are required to induce disease-associated spike signatures in the cerebellum and that spike signatures associated with ataxia, dystonia, and tremor can be generated by the same, otherwise healthy, cerebellar circuit.

Spike signatures in cerebellar nuclei neurons can be induced by specific stimulation paradigms of Purkinje cells.

A. Schematic of experimental setup with recordings in awake, head-fixed mice. B. Optopatcher recordings of cerebellar nuclei neurons. The opsin is expressed in Purkinje cells (pink) and recordings of nuclei neurons are performed. PC = Purkinje cell; CN = cerebellar nuclei; IO/CF = inferior olive/climbing fiber. C. Example validation that light stimulation of inhibitory Purkinje cells (blue bars) inhibits nuclei neurons during light stimulation. The lower trace is a blown-up view of the boxed area in the upper trace. D. Example of spike firing rate averaged over previous 50 ms at each occurring spike for the 5 s spike train in D’. D’’ 1 s spike train for the duration indicated in the square box in D’. Blue bars indicate light stimulation and are specific for ataxia, dystonia, and tremor (see methods for light stimulation parameters). All example traces originate from the same nuclei neuron, indicating that the cell’s spike signature can change depending on the light stimulation paradigm. E. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train, observe the shift in distribution from baseline during the different stimulation paradigms. F. Proportion of cells of each predicted spike signature during each of the light stimulations based on classifier model in Figure 1D. Control: n = 11 cells, N = 4 mice; ataxia: n = 9, N = 3; dystonia: n = 7, N = 3; tremor n = 9, N = 4. G. and H. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons.

Spike signatures can induce distinct motor phenotypes that mimic disease-related behaviors

During the head-fixed stimulation experiments, we observed subtle behavioral responses during unilateral cell-targeted optogenetic stimulation that suggested the different spike signatures might drive unique motor behaviors (Video 2). To explore this further in freely moving mice, we investigated whether the different spike signatures directly drive unique motor disturbances when a population of cells are induced to produce our defined spike signatures. To this end, L7Cre;R26ChR2mice were implanted with optical fibers that were bilaterally targeted to the interposed cerebellar nuclei (Figure 6A-C). This allowed for the targeting of Purkinje cell terminals in the cerebellar nuclei, as was done during our in vivo recordings in Figure 5, but with a larger population of cells affected by the stimulation. This also allowed the mice to move freely during stimulation, making a series of behavioral assays during stimulation possible (Figure 6D). Changes in behavior were immediate and severe as soon as the stimulation began. Each stimulation paradigm resulted in a different constellation of behaviors with the ataxia stimulation paradigm resulting in imprecise movements, the dystonia stimulation paradigm resulting in prominent dystonic posturing, and the tremor stimulation paradigm resulting in severe tremor (Video 3). Therefore, we assessed the mice for changes in gait, presence of dystonic movements, and severity of tremor. No single measurement can perfectly encapsulate the complexity of ataxia, dystonia, and tremor – all of which can appear in combination with the others (36–38). Additionally, each phenotype may affect the measurement of others. This was particularly apparent with measurements of gait. We found that all stimulation paradigms affected gait, resulting in visibly different foot placement (Figure 6E-F) and movement down the footprinting corridor (Video 4). Quantitatively, this was evident in shorter steps and less precision of fore and hindpaw placement (Figure 6I-K). We also considered the behavior of the mice in an open, flat, plexiglass arena for signs of dystonic movements (Figure 6L, Video 3). The dystonia stimulation paradigm resulted in the most frequent and strongest dystonic movements, while we also observed dystonic movements with the ataxia paradigm and abnormal movement – namely severe tremor – was noted with the tremor stimulation paradigm (Figure 6M). Mice were also placed in a custom-made tremor monitor (20) where they could freely ambulate during stimulation while an accelerometer mounted under the arena detected changes in acceleration (Figure 6D). While the ataxia stimulation paradigm did not significantly increase tremor from baseline, both the dystonia and tremor stimulation paradigms resulted in a significantly increased tremor at 10 Hz (Figure 6N-P). Though dystonia is often observed with tremor, it is possible that the detection of tremor in these animals was due to the jerkiness of their dystonic movements rather than an increase in smooth, rhythmic tremor that is more often associated with tremor disorders. Indeed, the tremor paradigm resulted in the most severe tremor of all stimulation conditions, producing a tremor severity that was more than ten-fold greater than that of the dystonia stimulation paradigm. Together, these measurements produced a complex representation of the underlying phenotypes. All three stimulation parameters resulted in a behavioral repertoire that was significantly different from baseline and, while there was some overlap of features, each stimulation paradigm produced the distinct and predicted respective motor phenotype (Figure 6Q-T). Together these data suggest that our classifier model’s defined spike signatures of disease-associated cerebellar nuclei spike trains are sufficient to produce the predicted corresponding abnormal motor phenotypes in mice.

Induced spike signatures elicit distinct cerebellar phenotypes.

A. Schematic of external view of bilateral optical fiber implant. B. Schematic of a coronal section from a mouse cerebellum with bilateral optic fiber implants directed towards the cerebellar nuclei. FN = fastigial nucleus. IN = interposed nucleus. DN = dentate nucleus. C. Photomicrograph of a Nissl-stained coronal section from a mouse cerebellum that had been implanted with optic fibers. Arrows = optic fiber tracks. Dotted lines surround the cerebellar nuclei indicated in B. Scale = 1 cm. D. Schematic of a mouse with bilateral optic fiber implants freely moving in a tremor monitor. E-K. Data associated with gait measurements. E-H. Example footprints from a single mouse before stimulation (baseline, E) and during ataxia (F), dystonia (G), and tremor stimulation (H). Scale = 1 cm. I-K. Measurements of gait including the length of the hindpaw stride (I), forepaw stride (J), and distance between the hind and forepaws (K). N = 8. * = p ≤ 0.05. ** = p ≤ 0.01. L. Example images of phenotypes associated with dystonia. 1 = erect tail. 2 = high stepping. 3 = kinked tail. 4 = hyperextension of the limbs. 5 = splayed toes. M. Dystonia rating of mice before stimulation and during stimulation with each paradigm. N = 8. * = p ≤ 0.05. N. Tremor signals detected via tremor monitor from a mouse before and during stimulation with each paradigm. Horizontal scale = 1 sec. Vertical scale = 50 mV. O. Population average power spectrums of tremor. Solid line = mean power. Shaded region = SEM. P. Peak tremor power of mice before and during stimulation with each paradigm. N = 8. * = p ≤ 0.05. ** = p ≤ 0.01. Q-S. 2-D comparisons of gait (hind to fore distance), dystonia (rating), and tremor (peak power) measurements from all mice. N = 8. T. 3-D plot of data in Q-S.

Discussion

In this study, we tested whether distinct spike train signatures in the interposed cerebellar nuclei explain why cerebellar dysfunction can cause multiple distinct motor impairments associated with movement disorders. By comparing spike activity across multiple mouse models of cerebellar disease, we found that the cerebellum can generate a range of dysfunctional spiking patterns. We found disease-specific spike train signatures using a classifier learner model, which allowed us to discover specific spike train parameters and their corresponding cutoff values that could distinguish the activity associated with these different disease states. When investigating whether these spike signatures are shared across mouse models with similar phenotypes due to different etiologies, we found that two types of spike train activity can cause ataxia, whereas specific spike train signatures are strongly associated with dystonia and tremor. We then tested whether we could optogenetically induce these signatures in an otherwise normal circuit. We found that the same neurons can generate healthy spike activity as well as a spectrum of disease-like spike activities. We then tested whether these optogenetically-induced disease-associated spike signatures could elicit similar behavioral abnormalities in the absence of any other primary genetic or circuit defects in awake and freely moving mice. The predominant behaviors that characterize each disease condition were successfully recapitulated. These data provide compelling evidence for the reliance of neurological phenotype presentation on the pattern of cerebellar circuit misfiring.

Several previous studies have proposed that distinct spike train patterns may correspond to distinct presentations of cerebellar disease (15, 29, 30, 39). Our work builds on these studies by quantitatively comparing spike train properties across, rather than within, mouse models. We demonstrate what aspects of the spike train patterns are distinct and shared across mouse models with different and similar disease phenotypes, respectively. We confirm prior research indicating the importance of spike train irregularity in disease presentation (15, 29, 39). We also show that spike train irregularity is insufficient to differentiate the spike train properties of dystonic and tremoring mice, which are differentiated from each other based on the instantaneous spike rate rather than pattern. Additionally, we confirm that the spike train signatures associated with different disorders can cause distinct disease phenotypes, thereby showing for the first time that distinct cerebellar spike train signatures are sufficient to drive a variety of motor impairments. Together, these data provide strong evidence that different spike train signatures do not only result from sensory feedback towards the cerebellum and that they may be a primary cause for motor impairments associated with cerebellar disease.

Our work suggests that there is a healthy range within the characteristics of cerebellar nuclei spiking activity and that cerebellar movement disorders are associated with a shift from this range in one or multiple features of spike train activity. We find some overlap and shared spike features between the different disease phenotypes and show that healthy cerebellar neurons can adapt multiple disease-associated spike train signatures. These findings suggest that pathophysiological spike train signatures are driven by a shift in neural function towards a disease state that can be independent of plasticity or circuit rewiring. Despite the dramatically different behavioral outcomes, the potential overlap and shared spike features, such as the irregular spike pattern found in the dystonia and tremor signatures, raise the strong possibility that co-expression or comorbidity of different motor disease behaviors may arise due to a spectrum of spike signal defects. This indicates that it is possible for neural signals to shift back and forth between healthy and disease states, potentially resulting in the transientness of behavioral impairments in certain cerebellar disorders. These findings also suggest that the most successful therapeutic avenues for cerebellar movement disorders should maintain cerebellar spiking activity within the healthy range without inadvertently inducing a different pathophysiological signature.

There has been (40), and there still remains (41), a great interest in understanding how altered cerebellar signals influence human behavior and disease. A pressing need to better define the architecture of these spiking abnomalities has arisen because of the success in using deep brain stimulation (DBS) and the potential for better tuning of the stimulation parameters for greater efficacy in treating different cerebellar-related disorders (ataxia: (42, 43); dystonia: (44); tremor: (45, 46); ataxia, dystonia, tremor: (41)). Recordings in human patients during DBS implantation have previously found physiological differences in the spike train patterns of basal ganglia neurons between dystonia and other movement disorders (for example; etiologically distinct dystonia: (47); dystonia with and without head tremor: (48); dystonia versus Parkinson’s disease: (49)). Our electrophysiological data confirm that differences in spike train properties are also found in the cerebellum and our optogenetic experiments suggest that these spike trains are not merely biomarkers that correlate with the disease state but are sufficient to cause motor impairments. Therefore, cerebellar-targeted DBS may be highly beneficial in normalizing the disease-causing spike train patterns.

A parallel motivation has fueled rodent studies to define the anatomical targets, stimulation parameters, and outcome efficiacy in detail (18, 25, 50–52). Although the exact mechanism(s) of DBS remains unclear (51, 53), there is strong support that at least one target of DBS is the actual neuronal signal itself (which the DBS may modify, enhance, dampen, and perhaps even interfere with the given neural spike defects). Indeed, the ability to alter cerebellar spike activity (54) has driven the investigation of the signal properties (40). Evidently, neurotransmission of the faulty activity patterns is required for the expression of disease behaviors (19). Recent works strongly support the hypothesis that changes in activity (rate, pattern, or both) could instigate a range of movement abnormalities (18, 20, 33, 55, 56). However, the current work defines the individual potential of these altered cerebellar signals to initiate specific motor changes. Dystonia can be reliably associated with irregular activity and tremor with highly rhthmic oscillations in the cerebellar nuclei neurons. Ataxia may be caused by a variety of changes that are different from the dystonia and tremor signatures in the severity of misfiring or proportion of neurons affected. Together, these data unveil a critical role for the cerebellum in triggering disease behaviors, with cauasative signals likely originating locally in its circuitry. Each disease may therefore arise as a result of a change in the balance and representaion of neural signatures.

Methods

Animals

Mice were housed in a Level 3, AALAS-certified vivarium. The Institutional Animal Care and Use Committee (IACUC) of Baylor College of Medicine (BCM) reviewed and approved all experimental procedures that involved mice. We used the following transgenic mouse lines: L7Cre (also known as Pcp2Cre) (57), Vgatfl(JAX:012897) (58), Ptf1aCre (JAX:023329) (59), Vglut2fl (JAX:012898) (60), Pdx1Cre (JAX:014647) (61), Car8wdl/wdl (JAX:004625) (10), Atxn1154Q (JAX:005601) (62), Ank1fl (JAX:036512) (28), and Ai32 (Rosa26lsl-ChR2-eYFP, JAX:024109) (63). We included mice from both sexes. We used ear punches from pre-weaned pups for PCR genotyping to identify the different transgenic alleles.

Headplate and craniotomy surgery for electrophysiology recordings

Prior to all recordings, we performed a surgery to attach a headplate over bregma and make a craniotomy over the interposed nucleus. This allowed us to collect stable recordings of cerebellar neuron activity while the mouse was awake and with or without a severe motor phenotype. This surgery was detailed in our previous publication (64). In short, mice were given preemptive analgesics including a subcutaneous injection of buprenorphine and meloxicam. Anesthesia was induced using inhaled isoflurane. During the surgery, the mice were kept on a heating blanket to maintain body temperature. Fur was removed from the surgery site using depilatory cream (Nair) and a small incision in the skin was made over the skull. Next, we used a dental drill to make a ∼2 mm diameter craniotomy over the interposed nucleus (6.4 mm posterior and 1.3 mm lateral to Bregma). The craniotomy was surrounded by a round chamber and filled with antibiotic ointment until the day of the recording. The recording chamber was closed with a screw top or silicone cap. We also placed a headplate with a hole over Bregma. A small piece of wire was placed vertically over Bregma to identify this anatomical marker on the day of recording. The recording chamber and headplate were attached to the skull using a combination of C and B Metabond Adhesive Luting Cement (Parkell) and methyl methacrylate dental cement (A-M Systems). Mice were allowed to recover from the surgery for at least three days before we began conducting the in vivo electrophysiological recordings.

Standard in vivo electrophysiological recordings in awake mice

We performed in vivo electrophysiology recordings according to experimental procedures detailed in our previous publications (20, 64, 65). In brief, we placed mice on a rotating foam wheel and stabilized their heads by screwing the implanted headplates to the recording rig. We removed the antibiotic ointment from the recording chamber and replaced it with sterile physiological saline solution (unless otherwise specified). We measured the coordinates of Bregma to determine the coordinates where the electrode would penetrate the surface of the cerebellum. We used tungsten electrodes (2-8 MΩ, Thomas Recording) attached to a preamplifier head stage (NPI Electronic instruments) for our recordings. The position of the recording electrode was controlled using a motorized micromanipulator (MP-255; Sutter Instrument Co.). Neural signals were amplified and bandpass filtered (0.3-13 kHz) using an ELC-03XS amplifier (NPI Electronics) before being digitized using a CED board. All signals were recorded using Spike2 software (CED). We only included neurons recorded between 2.5 and 3.5 mm from the surface that did not exhibit complex spikes (characteristic for Purkinje cell firing) in our analyses of cerebellar nuclei neuron firing patterns. This paper includes reanalyzed data from previously reported studies (18, 20, 28, 65).

Spike sorting and analyses

We previously showed that recording duration influences the estimation of neural firing parameters (66). Therefore, all parameter estimates in this paper are based on 30 seconds (s)-long recordings. We used Spike2 software to sort out spikes from electrophysiological recordings. Information about spike timing were imported and saved in MATLAB databases using custom written code. We described the spike analyses using twelve parameters investigating the properties of interspike intervals (ISI) within the recording; (1) ; (2) mean instantaneous firing rate = mean (ISI−1); (3) median instantaeous firing rate = median (ISI−1); (4) ; (5) ; (6) skewness = median instantaneous firing rate - firing rate; (7) ISI25 = % ISI > 25 ms; (8) ISI100 = % ISI > 100 ms; (9) inter burst pause = mean (ISI > (5 * mean (ISI)); (10) kurtosis = % ISI at model(ISI); (11) ; (12) oscillation pearks = number of peaks ″a″. For the calculation of 11 and 12, we performed an autocorrelation analysis for all spikes in the 30 s recording, calculated the rhythmicity index, and found oscillation peaks as previously described (67, 68). We used a bin width of 5 ms. The first oscillation peak was determined as the highest bin between 10 ms and 1.5 times the mean ISI for a given neuron. The difference between the baseline level and the height of the peak was denoted as a1, the difference between baseline and the depth of the trough was denoted as b1, and z was the difference between baseline and the total number of spikes included in the analyses. Each subsequent peak was determined as the highest bin between the delay-time of the previous trough and an+a1+10 ms, where an is the time of the previous peak. The first trough was determined as the lowest bin between the first peak (a1) and an + a1. Peaks and troughs were only accepted if their sum was higher than four times the standard deviation of autocorrelation between 0.96 and 1 s lag-time, or if the peak was higher than baseline plus two times the standard deviation and the trough was lower than baseline minus two times the standard deviation.

Supervised classifier learner

We trained our classifier learner using cerebellar nuclei neuron recordings from mixed background control mice (control), L7Cre;Vgatfl/fl mice (ataxia), Ptf1aCre;Vglut2fl/flmice (dystonia), and harmaline-injected mice (tremor). We have previously tested these models and consider them as reliable archetypical models for cerebellar movement disorders (17, 18, 20). For the analyses, we reanalyzed previous recordings (18, 20, 28, 35, 65) in addition to newly acquired ones. We only included recordings with a duration of 30 s or longer and calculated parameter estimates as described above. If recordings were longer than 60 s, we also calculated parameter estimates for the second 30 s period. We randomly assigned the parameter estimates calculated from the first 30 s of the recording to the training or validation data set and, when available, we assigned the parameter estimates from the second 30 s of the recording to the remaining data set. This resulted in a training set comprised of 29 neurons from 9 control mice, 19 neurons from 5 ataxic mice, 19 neurons from 8 dystonic mice, and 21 neurons from 6 tremoring mice. The validation set contained 30 neurons from 9 control mice, 18 neurons from 5 ataxic mice, 17 neurons from 7 dystonic mice, and 18 neurons from 5 tremoring mice.

We used the built-in supervised machine learning, MATLAB Classification Learner APP (The Mathworks Inc, version R2021a) to define spike signature parameter value cutoffs. We imported the parameters describing the spiking activity from an Excel (Microsoft) worksheet. We used “Group” (control, ataxia, dystonia, or tremor) as the “Response Variable” and the 12 parameters described above as the “Predictor Variables.” We trained a “Coarse Tree” with maximum number of splits = 3, split criterion = Gini’s diversity index, and surrogate decision splits = off. We exported the trained model to the workspace to validate the classifier learner. We assigned signature identity to recordings based on the trained classifier tree.

Optopatcher in vivo electrophysiology recordings in awake mice

Optopatcher experiments were performed as previously described (69). We prepared the mice for recording by performing a headplate and craniotomy surgery as described above. The electrophysiology rig and setup were the same as described above with the following differences: We used an optopatcher device (ALA Scientific Instruments Inc) that allows for the placement of a custom-made optical fiber (Thorlabs, #FT200UMT) within a glass recording electrode (Harvard Apparatus, #30-0057). The tip of the optical fiber was placed near the tip of the recording electrode and was illuminated via a 465 nm LED light source (ALA Scientific Instruments Inc). On the day of the recording, we pulled glass electrodes (Sutter Instruments, #P-1000), filled the electrodes with physiological saline, and measured their impedance using an ELC-03XS amplifier (NPI Electronics) before recording. Only electrodes with 2-15 MΩ impedance were used. Light stimulation was triggered using custom Spike2 scripts paired with a CED board (CED). All optopatcher recordings were performed in L7Cre;Ai32 mice that express channelrhodopsin in Purkinje cells. Nuclei neurons included in our analyses were cell recordings between 2.5-3.5 mm from the surface of the cerebellum and were inhibited by brief light stimulation at 465 nm. After we found and isolated a cell, we slowly ramped up the brightness of this brief stimulation until we found the minimal light power that modulated the spiking activity. This minimal power was then used to stimulate the cell with the various test parameters during the recording session.

Optogenetic light stimulation paradigms for optopatcher recordings and behavioral assays

We used the following stimulation parameters to measure signature-specific responses; ataxia: 50 Hz (10 ms light on/ 10 ms light off) square pulses; dystonia: at least 500 ms of 50 Hz (10 ms light on/ 10 ms light off) square pulses interspersed with at 75% chance of 250 ms pauses in stimulation; tremor: 100 Hz pulses (5 ms light on/ 5 ms light off) overlaid with a 10 Hz sinusoid (50 ms parabolic increase and decrease of light power followed by 50 ms light off).

Bilateral optic fiber implant surgery for in vivo behavioral assays

We implanted mice with optical fibers (ThorLabs, #FT200UMT) targeted bilaterally to the interposed cerebellar nuclei to assess the motor phenotypes that result from our stimulation parameters. We previously described this surgical procedure (Brown et al., 2020). Briefly, the mice were given subcutaneous injections of sustained-release buprenorphine and meloxicam as preemptive analgesics. Anesthesia was induced with 3% isoflurane gas and the anesthetic plane was maintained with 2% isoflurane gas. The mice were placed on a heating blanket and the head was stabilized in a stereotaxic frame (David Kopf Instruments). Fur was removed from the top of the head using depilatory cream (Nair). The surgical site was cleaned using alternating applications of betadine scrub and alcohol. An incision was made in the skin to expose the skull from anterior to bregma to the posterior aspect of the occipital plate. The fascia was removed from the top of the skull and a scalpel was used to etch fine grooves into the top of the skull. A small craniotomy was made in the parietal plate with a dental drill in order to attach a skull screw (00-90 x 1/16 flat point stainless steel machine screw) to anchor the implant to the skull. The skull screw was advanced only until the point that it was stable in the skull and care was taken to ensure it did not contact the brain. The implant sites were determined by measuring 6.4 mm posterior and ±1.5 mm lateral to bregma. Small craniotomies were made at these sites using a dental drill. The base of the fibers were placed on the surface of the cerebellum and were advanced ventrally for 2 mm. A small amount of antibiotic ointment was placed around each fiber to prevent Metabond or dental cement from entering the craniotomy. Metabond was applied over the entire exposed area of skull and around the skull screw and fibers. Two short segments of wire were embedded in the Metabond to allow the experimenter to hold the mouse’s head while attaching and removing the fiber patch cables from the implant. Dental cement was placed over the Metabond. The mice were allowed to recover for at least 3 days before any behavioral assessments were made. Subcutaneous injections of meloxicam were provided every 24 h during the recovery period. Eight mice (number of mice, N) were implanted and tested with all three behavioral assays.

Gait analysis

Measurements of gait were made for all mice both before (baseline) and during each stimulation parameter. To do this, the mice were briefly scruffed and then contrasting colored non-toxic paints were applied to the soles of their forepaws and hindpaws (blue paint for forepaws and red paint for hindpaws). The mice were then gently set down on a piece of blank white paper positioned between two parallel plexiglass barriers with a dark enclosed area at the end of the corridor. Adult mice are naturally inclined to walk towards the dark enclosed area, meanwhile depositing paint on the paper with each footstep. The stimulation, if present, was extinguished once the mice entered the enclosed area. The mice were allowed to rest for at least 2 min between each trial. A trial was considered successful if there were at least four hind and forepaw prints for both the left and right feet that were visible, steady (the mouse was neither running or stopping midway through the series of footsteps), and in as straight of a line as possible (the mouse was not actively turning during the trial). Optic patch cables were connected to the implant during every trial (including baseline runs when stimulation was not present). At least two successful trails were collected per stimulation parameter. For analysis, measurements from two trials of the same type were averaged to determine the values for each mouse. These measurements were stride, the distance from one footprint to the next from the same foot and hind to fore distance, and hind-to-fore distance, the distance from one hindpaw placement from a corresponding forepaw placement on the same side of the body. Three of each measurement were made per footprint and were averaged to determine the measurements per foot, per trial. These trials were then averaged to determine the final measurements for each mouse. A repeated-measures one-way ANOVA with a Tukey multiple comparison adjustment was performed to determine significance. Significance values were indicated as not significant (ns) if p > 0.05, * = p ≤ 0.05. ** = p ≤ 0.01.

Dystonia rating scale

Mice were placed in a rectangular plexiglass arena in order to phenotypically rate the frequency and severity of dystonic behavior as described previously (Jinnah et al., 2000; Pizoli et al., 2002). The videos were captured to include the animals’ behavior before (baseline) and during each stimulation paradigm (ataxia, dystonia, and tremor). Each stimulation period lasted 2 min, during which the mouse was allowed to ambulate freely as well as in response to disturbance by the experimenter. The rating period excluded the first 10 s after stimulation was initiated to avoid including behavior in response to the sudden application of stimulation. Mice were given a score of 0 if no motor abnormalities were identified, 1 if there was abnormal motor behavior that was not obviously dystonic (such as severe tremor), 2 if there was a mild motor phenotype that included dystonic behaviors that may occur only in response to being disturbed by the experimenter or unprovoked (such as brief hyperextension of the limbs or extension of the neck and/or back), 3 if there was moderate severity or frequent unprovoked dystonic behaviors, 4 if there were severe unprovoked dystonic behaviors, and 5 if there were severe unprovoked dystonic behaviors that made goal-directed ambulation extremely difficult or impossible for an extended time. A detailed description of dystonic phenotypes in mice can be found in Brown et al. 2022 (22). Wilcoxon matched-pairs signed rank tests were performed with a post-hoc Holm-Sidak method for p-value adjustment in order to determine significance. Significance values were indicated as not significant (ns) if p > 0.05, * = p ≤ 0.05. ** = p ≤ 0.01.

Tremor analysis

The amplitude and frequency of tremor before (baseline) and during optogenetic stimulation was measured using a custom-made tremor monitor as previously described (20, 35, 65)The mice were implanted with optical fibers targeted to the interposed nucleus as described above. LED drivers were connected to the implant via optical patch cables and placed above the tremor monitor chamber. The tremor monitor chamber is a clear plexiglass box that is suspended by elastic cords that are connected to each corner of the box. An accelerometer is securely mounted to the bottom of the box to detect the tremor of the mouse within the box. The mice were placed in the chamber and were able to freely ambulate while in the box. They were given 2 min to acclimate to the novel tremor monitor environment before 2 min duration baseline recordings were made. The mice were then stimulated with the ataxia, dystonia, and tremor stimulation parameters for 2 min per stimulation period. At least 2 min were allowed to elapse between stimulation periods. Mice were encouraged to ambulate a similar amount during all recording periods. The output from the tremor monitor was amplified and lowpass filtered at 5 kHz (Brownlee Precision, Model 410) before being digitized by a CED board and analyzed using Spike2 scripts. Tremor monitor recordings were centered on zero and downsampled using the Spike2 interpolate function. A power spectrum analysis with a Hanning window was performed on each recording period. The peak power was calculated as the maximum power between 0 and 30 Hz, which is consistent with the range that we typically observe for physiological and pathological tremor in mice (20). A repeated-measures one-way ANOVA with a Tukey multiple comparison adjustment was performed to determine significance. Significance values were indicated as not significant (ns) if p > 0.05, * = p ≤ 0.05. ** = p ≤ 0.01.

Histology

After electrophysiology and behavior experimentation, mice were heavily anesthetized with avertin (2,2,2-tribromoethanol, Sigma-Aldrich, St. Louis, MO, USA; #T48402) and transcardially perfused with ice-cold phosphate buffered saline (PBS, 1x) followed by ice-cold 4% paraformaldehyde (PFA). The implants were removed, and then the brains were dissected out of the skull. The brains were post-fixed in PFA at 4°C for at least 24 hours. They were then cryoprotected via a sucrose gradient, starting at 15% sucrose in PBS followed by 30% sucrose in PBS. The brains were imbedded in Tissue-Tek O.C.T. Compound (Sakura, Torrance, CA, USA), frozen, and then sliced on a cryostat to produce 40 µm thickness sections. The tissue sections were washed in PBS, mounted onto slides, and allowed to dry on the slide for at least 2 hours. The sections were put in a xylene and ethanol series with ∼2 min per submersion in the following order: 100% xylene, 100% xylene, 100% ethanol, 100% ethanol, 90% ethanol, 70% ethanol. The sections were then placed in water for ∼2 min and then submerged in cresyl violet solution until the stain was sufficiently dark. They were then dehydrated by reversing the series with ∼30 s per submersion. Finally, Cytoseal XYL mounting media (Thermo Scientific, Waltham, MA, USA, #22-050-262) and a cover slip were placed on the slides. The slides were allowed to dry in a fume hood before imaging. Photomicrographs were obtained using a Leica DM4000 B LED microscope equipped with a Leica DMC 2900 camera and Leica Application Suite X (LAS X) software. Images were corrected for brightness and contrast using Adobe Photoshop (Adobe Systems, San Jose, CA, USA). Figure panels were made using Adobe Illustrator software.

Supporting information

eLife Assessment

This is an important study because it provides evidence that specific neuronal firing patterns in deep cerebellar nuclei map onto specific behavioral movement disorder phenotypes. The optogenetic manipulations and resulting neuronal and behavioral outcomes are highly compelling, but the development of the classifier tool was incomplete. This study contributes to the fields of cerebellar physiology and movement disorders because it puts forth a map of relationships between neuronal firing patterns and multiple distinct movement phenomena, providing a comprehensive view that goes beyond most studies which typically examine one phenomenon in isolation.

Reviewer #1 (Public Review):

Summary:

Van der Heijden et al perform an ambitious analysis of single-unit activity in the interposed nuclei of multiple mouse models of cerebellar dysfunction. Based on these recordings, they develop a classifier to predict the behavioral phenotype (ataxic, dystonic, or tremor) of each model, suggesting that highly regular spiking is associated with ataxia, irregular spiking is associated with dystonia and rhythmic spiking is associated with tremor. After developing this classifier, they show that activating Purkinje neurons in different patterns that evoke interposed nuclear activity similar to their "ataxic", "dystonic", and "tremor" firing patterns induce similar behaviors in healthy mice. These results show convincingly that specific patterns of cerebellar output are sufficient to cause specific movement abnormalities. The extent to which cerebellar nuclear firing patterns are solely responsible for phenotypes in human disease remains to be established, however.

Strengths:

Major strengths are the recordings across multiple phenotypic models including genetic and pharmacologic manipulations, and the robust phenotypes elicited by Purkinje neuron stimulation.

Weaknesses:

The number of units recorded was small for each model (on the order of 20), limiting the conclusions that can be drawn from the recording/classifier experiments.

Reviewer #2 (Public Review):

Cerebellar diseases can manifest as various behavioral phenotypes, such as ataxia, dystonia, and tremor. In this study, Heijden and colleagues aim to understand whether these differing behavioral phenotypes are associated with disease-specific changes in the firing patterns of cerebellar output neurons in the cerebellar nuclei (CN). The authors effectively demonstrate that across different mouse models of cerebellar disease, there are distinct changes in the firing properties of CN neurons. They take a crucial step further by attempting to replicate disease-specific firing patterns in the cerebellar output neurons of healthy (control) mice using optogenetics. When Purkinje cells are stimulated in a manner that results in similar firing properties in CN neurons, the authors observe a variety of atypical behavioral responses, many of which align with the behavioral phenotypes observed in mouse models of the respective diseases.

Overall, the primary results are quite convincing. Specifically, they show that (1) different mouse models of cerebellar disease exhibit different statistics of firing in CN neurons, and (2) driving CN neurons in a time-varying manner that mimics the statistics measured in disease models results in behavioral phenomena reminiscent of the disease states. These findings suggest that aberrant activity in the CN can originate from various sources (e.g., developmental circuit deficits, abnormal plasticity, insult), but ultimately, these changes are funneled through the CN neurons, whose firing rates are affected, and this, in turn, drives aberrant behavior. This is a noteworthy observation that underscores the potential of targeting these output neurons in the treatment of cerebellar disease. Moreover, this manuscript provides valuable insights into the firing patterns associated with the most common cerebellar-dependent disease phenotypes.

However, the paper falls short in terms of the classifier model itself. The current implementation of this classifier appears to be rather weak. For instance, the cross-validated performance on the same disease line mouse model for tremor is only 56%. While I understand that the classifier aims to simplify a high-dimensional dataset into a more manageable decision tree, its rather poor performance undermines the authors' main objectives. In a similar vein, although focusing on three primary features of spiking statistics identified by the decision tree model (CV, CV2, and median ISI) is useful for understanding the primary differences between the firing statistics of different mouse models, it results in an overly simplistic view of this complex data. The classifier and its reliance on the reduced feature set are the weakest points of the paper and could benefit from further analysis and a different classification architecture. Nevertheless, it is commendable that the authors have collected high-quality data to validate their classifier. Particularly impressive is their inclusion of data from multiple mouse models of ataxia, dystonia, and tremor, enabling a true test of the classifier's generalizability.

Reviewer #3 (Public Review):

Summary:

This manuscript looks at the single-cell spike signatures taken from in vivo cerebellar nuclear neurons from awake mice suffering from 3 distinct diseases and uses a sophisticated classifier model to predict disease based on a number of different parameters about the spiking patterns, rather than just one or two. Single read-outs of spike firing patterns did not show significant differences between all 4 groups meaning that you need to analyze multiple parameters of the spike trains to get this information. The results are really satisfying and intriguing, with some diseases separating very well, and others having more overlap. It also represents a significant advancement for the rigor and creativity used for analyzing cerebellar output spike patterns. I really like this paper, it's a clever idea and has been done very well.

The authors examine multiple distinct forms of different diseases, including different types of ataxia, dystonia, and tremor. While some of the interpretation of this work remains unclear to this reviewer (in particular Figure 2, with ataxia models), I applaud the rigor and sharing of complex data that is not always straightforward to understand.

Strengths:

The work is technically impressive and the analysis pushes the envelope of how cerebellar dysfunction is classified, which makes it an important paper for the field. It's well written. The approach it is taking is clever. The analysis is thorough, and the authors examine a wide array of different disease models, which is time-consuming, costly, and very challenging to do. It's a very strong manuscript.

Weaknesses:

Weaknesses are few and quite minor. Some rewriting could be done to make certain sections clearer.

Author response:

The following is the authors’ response to the current reviews.

Reviewer #1 (Recommendations For The Authors):

I still find it really impressive that the Purkinje cell stimulation so closely mimics the pathogenic phenotypes - in my opinion, the strongest part of the paper. I would like just a little clarification on some of my previous questions.

Major points:

(1) Can the authors clarify where the new units came from? Are these units that were recorded before the initial submission and excluded, but are now included? If so, why were they excluded before? Or are these units that were recorded since the original submission?

The number of units increased in Figure 1 for three reasons: 1) We have now plotted the classifier results in Figure 1 instead of the validation results, which have been moved to Figure 1 Supplement 3. 2) In response to reviewer comments, we no longer include units that had >60 s of recording in both our model creation and validation. We had previously used 30 s for creating the model and a different 30 s for validating the model, if an additional 30 s were available. 3) We changed our model creation and validation strategy based on previous reviewer comments. The new units in Figures 2-4 were taken from our pool of previously collected but unanalyzed data (we collect neural data on a rolling basis and thus these data were not initially available). We were fortunate to have these data to analyze in order to address the concerns about the number of cells included in the manuscript. The number of units increased in Figure 5 because new units were recorded in response to reviewer comments.

(2) Why did some of the neuron counts go down? For example, in Pdx1Cre;Vglut2fl/fl mice, the fraction of units with the control signature went from 11/21 to 7/23. Is this because the classifier changed between the original submission and the revision?

Yes, the proportion of cells matching each classification changed due to the different parameters and thresholds used in the updated classifier model.

Minor points:

In the Discussion: "We find some overlap and shared spike features between the different disease phenotypes and show that healthy cerebellar neurons can adapt multiple disease-associated spike train signatures." I think "adapt" should be "adopt"

In the Discussion: "compare" is misspelled as "compared"

Thank you for bringing these typos to our attention. We will upload a new version of the text with the typos corrected.

The following is the authors’ response to the original reviews.

We would like to thank the Reviewers for providing excellent and constructive suggestions that have enabled us to strengthen our overall presentation of our data. We have addressed each of the comments by altering the text, providing additional data, and revising the figures, as requested.

Below are our explanations for how we have altered the manuscript in this revised version.

Recommendations for the authors:

I think you will have seen from the comments that there was great enthusiasm for the importance of this study. There were also shared concerns about how the classifier may be inadequate in its current format, as well as specific suggestions to consider to improve. I hope that you will consider a revision to really amplify the impact of the importance of this study.

Reviewer #1 (Recommendations For The Authors):

Distinct motor phenotypes are reflected in different neuronal firing patterns at different loci in motor circuits. However, it is difficult to determine if these altered firing patterns: 1) reflect the underlying neuropathology or phenotype, 2) whether these changes are intrinsic to the local cell population or caused by larger network changes, and 3) whether abnormal firing patterns cause or reflect abnormal movement patterns. This manuscript attempts to address these questions by recording neural firing patterns in deep cerebellar nucleus neurons in several models of cerebellar dysfunction with distinct phenotypes. They develop a classifier based on parameters of single unit spike trains that seems to do an inconsistent job of predicting phenotype (though it does fairly well for tremor). The major limitation of the recording/classifier experiments is the low number of single units recorded in each model, greatly limiting statistical power. However, the authors go on to show that specific patterns of Purkinje cell stimulation cause consistent changes in interposed nucleus activity that map remarkably well onto behavioral phenotypes. Overall, I did not find the recording/classifier results to be very convincing, while the stimulation results strongly indicate that interposed nucleus firing patterns are sufficient to drive distinct behavioral phenotypes.

We thank the reviewer for their comments. We describe below how we have addressed the major concerns.

Major concerns:

(1) I don't think it's legitimate to use two 30-second samples from the same recording to train and validate the classifier. I would expect recordings from the same mouse, let alone the same unit, to be highly correlated with each other and therefore overestimate the accuracy of the classifier. How many of the recordings in the training and validation sets were the same unit recorded at two different times?

We previously published a paper wherein we measured the correlation (or variability) between units recorded from the same mouse versus units recorded from different mice (see: Van der Heijden et al., 2022 – iScience, PMID: 36388953). In this paper we did not find that nuclei neuron recordings from the same mouse were more correlated or similar to each other than recordings from different mice.

Upon this reviewer comment, however, we did observe strong correlations between the two 30-second samples from the same recording units. We therefore decided to no longer validate our classifier based on a training and validation sets that had overlapping units. Instead, we generated 12 training sets and 12 non-overlapping validation sets based on our entire database. We then trained 12 classifier models and ranked these based on their classification ability on the validation sets (Figure 1 – supplemental Figure 3). We found that the top two performing classifier models were the same, and used this model for the remainder of the paper.

(2) The n's are not convincing for the spike signature analyses in different phenotypic models. For example, the claim is that Pdx1Cre;Vglut2fl/fl mice have more "control" neurons than ouabain infusion mice (more severe phenotype). However, the numbers are 11/21 and 7/20, respectively. The next claim is that 9/21 dystonic neurons are less than 11/20 dystonic neurons. A z-test for proportions gives a p-value of 0.26 for the first comparison and a pvalue of 0.44 for the second. I do not think any conclusions can be drawn based on these data.

We included more cells in our analyses and found that the z-test for n the proportion of cells with the “control” and “dystonia” signature is indeed statistically significant.

(3) Since the spiking pattern does not appear to predict an ataxic phenotype and the n's are too small to draw a conclusion for the dystonic mice, I think the title is very misleading - it does not appear to be true that "Neural spiking patterns predict behavioral phenotypes...", at least in these models.

We have changed the title to: “Cerebellar nuclei cells produce distinct pathogenic spike signatures in mouse models of ataxia, dystonia, and tremor.” We feel that this new title captures the idea that we find differences between spike signatures associated with ataxia, dystonia, and tremor and that these signatures induce pathological movements.

(4) I don't think it can be concluded from the optogenetic experiments that the spike train signatures do not depend on "developmental changes, ...the effect of transgene expression, ... or drug effects outside the cerebellum." The optogenetic experiments demonstrate that modulating Purkinje cell activity is sufficient to cause changes in DCN firing patterns and phenotypes (i.e., proof-of-principle). However, they do not prove that this is why DCN firing is abnormal in each model individually.

Thank you for highlighting this section of the text. We agree that the optogenetic experiments cannot explain why the DCN is firing abnormally in each model. We have edited this section of the text to prevent this conclusion from being drawn by the readers.

Minor points:

(1) It would be nice to see neural recordings in the interposed nucleus during Purkinje terminal stimulation to verify that the firing patterns observed during direct Purkinje neuron illumination are reproduced with terminal activation. This should be the case, but I'm not 100% certain it is.

We have edited the text to clarify that representative traces and analysis of interposed nucleus neurons in response to Purkinje terminal stimulation are the data in Figure 5.

(2) How does the classifier validation (Fig. 1E) compare to chance? If I understand correctly, 24/30 neurons recorded in control mice are predicted to have come from control mice (for example). This seems fairly high, but it is hard to know how impressive this is. One approach would be to repeat the analysis many (1000s) of times with each recording randomly assigned to one of the four groups and see what the distribution of "correct" predictions is for each category, which can be compared against the actual outcome.

We have now also included the proportion of spike signatures in the entire population of neurons and show that the spike signatures are enriched in each of the four groups (control, ataxia, dystonia, tremor) relative to the presence of these signatures in the population (Figure 1E).

(3) I don't think this is absolutely necessary, but do the authors have ideas about how their identified firing patterns might lead to each of these phenotypes? Are there testable hypotheses for how different phenotypes caused by their stimulation paradigms arise at a network level?

We have added some ideas about how these spike signatures might lead to their associated phenotypes to the discussion.

Reviewer #2 (Recommendations For The Authors):

(1) As mentioned earlier, my main concern pertains to the overall architecture and training of the classifier. Based on my reading of the methods and the documentation for the classifier model, I believe that the classifier boundaries may be biased by the unequal distribution of neurons across cerebellar disease groups (e.g., n=29 neurons in control versus n=19 in ataxics). As the classifier is trained to minimize the classification error across the entire sample, the actual thresholds on the parameters of interest may be influenced by the overrepresentation of neurons from control mice. To address this issue, one possible solution would be to reweight each group so that the overall weight across classes is equal. However, I suggest a better strategy might be to revise the classifier architecture altogether (as detailed below).

We have retrained the classifier model based on equal numbers of ataxic, dystonic, and tremor cells (n=20) but we intentionally included more control cells (n=25). We included more control cells because we assume this is the baseline status for all cerebellar neurons and wanted to avoid assigning disease signatures to healthy neurons too easily.

(2) As the authors make abundantly clear, one mouse model of disease could potentially exhibit multiple phenotypes (e.g., a mouse with both ataxia and tremor). To address this complexity, it might be more valuable to predict the probability of a certain CN recording producing specific behavioral phenotypes. In this revised approach, the output of the classifier wouldn't be a single classification (e.g., "this is an ataxic mouse") but rather the probability of a certain neural recording corresponding to ataxia-like symptoms (e.g., "the classifier suggests that this mouse has a 76% likelihood of exhibiting ataxic symptoms given this CN recording"). This modification wouldn't require additional data collection, and the exemplar disease models could still be used to train such a revised network/classifier, with each mouse model corresponding to 0% probability of observing all other behavioral phenotypes except for the specific output corresponding to the disease state (e.g., L7CreVgat-fl/fl would be 0% for all categories except ataxia, which would be trained to produce a score of 100%). This approach could enhance the validation results across other mouse models by allowing flexibility in a particular spike train parameter to produce a diverse set of phenotypes.

This is a great comment. Unfortunately, our current dataset is constrained to fully address this comment for the following reasons:

- We have a limited number of neurons on which we can train our classifier neurons. Further dividing up the groups of neurons or complicating the model limited the power of our analyses and resulted in overfitting of the model on too few neurons.

- The recording durations (30 seconds) used to train our model are likely too short to find multiple disease signatures within a single recording. We feel that the complex phenotypes are likely resulting from cells within one mouse exhibiting a mix of disease signatures (as in the Car8wdl/wdl mice).

We think this question would be great for a follow-up study that uses a large number of recordings from single mice to fully predict the mouse phenotype based on the population spike signatures.

To limit confusion about our classifier model, we have also altered the language of our manuscript and refer to the cells exhibiting a spike signature instead of predicting a phenotype.

However, the paper falls short in terms of the classifier model itself. The current implementation of this classifier appears to be rather weak. For instance, the crossvalidated performance on the same disease line mouse model for tremor is only 56%. While I understand that the classifier aims to simplify a high-dimensional dataset into a more manageable decision tree, its rather poor performance undermines the authors' main objectives. In a similar vein, although focusing on three primary features of spiking statistics identified by the decision tree model (CV, CV2, and median ISI) is useful for understanding the primary differences between the firing statistics of different mouse models, it results in an overly simplistic view of this complex data. The classifier and its reliance on the reduced feature set are the weakest points of the paper and could benefit from further analysis and a different classification architecture. Nevertheless, it is commendable that the authors have collected high-quality data to validate their classifier. Particularly impressive is their inclusion of data from multiple mouse models of ataxia, dystonia, and tremor, enabling a true test of the classifier's generalizability.

We intentionally simplified our parameter space from a high-dimensional dataset into a more manageable decision tree. We did this for the following reasons:

- The parameters, even though all measuring different features, are highly correlated (see Figure 1 – supplemental Figure 2). Further, we were training our dataset on a limited number of recordings. We found that including all parameters (for example using a linear model) caused overfitting of the data and poor model performance.

- Describing the spike signatures using a lower number of parameters allowed us to design optogenetic parameters that would mimic this parameter space. This would be infinitely more complex with a bigger parameter space.

We agree with the reviewer that inclusion of multiple mouse models in addition to the optogenetics experiments provide the classifier’s generalizability.

Minor Comments:

(1) The blown-up CN voltage traces in Figures 5C and Supplementary Figure 2B appear more like bar plots than voltage traces on my machine.

Thank you for bringing this to our attention. We have improved the rendering of the traces.

(2) The logic in lines 224-228 is somewhat confusing. The spike train signatures are undoubtedly affected by all the factors mentioned by the authors. What, I believe, the authors intend to convey is that because changes in CN firing rates can be driven by multiple factors, it is the CN firing properties themselves that likely drive disease-specific phenotypes.

We agree that our discussion of the CN firing needs clarification. We have made the appropriate edits in the text.

Reviewer #3 (Recommendations For The Authors):

It's quite astounding that this can be done from single spike trains from what are almost certainly mixed populations of neurons. Could you add something to the discussion about this? Some questions that could be addressed would be would multiple simultaneous recordings additionally help classify these diseases, or would non-simultaneous recordings from the same animal be useful? Also more discussion about which cells you are likely recording from would be useful.

Thank you for this suggestion. We have added discussion about multiple recordings, simultaneous vs non-simultaneous recordings, and our thoughts on the cell population recorded in this work.

Data in figure 2 is difficult to understand - it appears that the majority of dysregulated cells in 2 ataxic models are classified as dystonia cells, not ataxic cells. This appears surprising as it seems to be at odds with earlier data from Fig 1. In my opinion, it is not discussed adequately in the Results or Discussion section.

We have added further discussion of the ataxia models represented in Figures 1 and 2.

Minor comment:

The colours of the subdivisions of the bars in 2C and 3C, and the rest of the paper appear to be related to the groups in the middle (under "predicted"), but the colours are much paler in the figure than in the legend, although the colours in the bars and the legends match in the first figure (1E). Does this signify something?

These figures were remade with the same colors across the board.

Version 2

Abstract

Abstract

The cerebellum contributes to a diverse array of motor conditions including ataxia, dystonia, and tremor. The neural substrates that encode this diversity are unclear. Here, we tested whether the neural spike activity of cerebellar output neurons is distinct between movement disorders with different impairments, generalizable across movement disorders with similar impairments, and capable of causing distinct movement impairments. Using in vivo awake recordings as input data, we trained a supervised classifier model to differentiate the spike parameters between mouse models for ataxia, dystonia, and tremor. The classifier model correctly assigned mouse phenotypes based on single neuron signatures. Spike signatures were shared across etiologically distinct but phenotypically similar disease models. Mimicking these pathophysiological spike signatures with optogenetics induced the predicted motor impairments in otherwise healthy mice. These data show that distinct spike signatures promote the behavioral presentation of cerebellar diseases.

Introduction

There exists a historical and rich curiosity in understanding the role of the cerebellum in movement, dating back to the pioneering work of Flourens (1), with an equally long interest in investigating how it alters movement (2,3). From these earlier studies it was clear that a defective cerebellum causes a range of devastating problems in the ability to control voluntary, intentional actions including coordination, posture, and balance. Accordingly, in humans, the equivalent defects had clear pathophysiological consequences. Patients with various cerebellar lesions showed loss of precise motor coordination, which, as the consequence of disease, is referred to as ataxia; whereas some subjects displayed overt and sometimes exaggerated oscillatory movements, consistent with tremor (2–4). In other cases, the examiner would report uncontrolled muscle contractions, which is now a contributing feature of dystonia (5). These classic observations underscored the heterogeneity of motor disturbances caused by cerebellar dysfunction. But, even then, the question of how the cerebellum creates such behavioral diversity was already imminent. Despite major advances in understanding the basic anatomy and circuitry of the cerebellum, it is still unclear how disease behaviors emerge of these circuits. In this regard, the specific substrates that underlie each disorder could hold the key for improving the design of therapies and treatments.

Consistent with the outcomes of removing or lesioning the cerebellum in pigeons, dogs, monkeys, and humans (2–4), the heterogeneity of cerebellar movement disturbances was later confirmed in genetic mouse models. Much of the current excitement about the cerebellum spawned from these initial genetic models because of the overt motor disturbances that were caused by spontaneous mutations in genes that are now known to be associated with cerebellar development or cerebellar degeneration (6,7). Prior to genetic sequencing though, mutant mice such as hot-foot (8), weaver (9), trembler (6), waddles (10), staggerer (11), stumbler (12), tottering (13), and lurcher (14) were named according to phenotype of their abnormal movements, which are as diverse as their names imply. Modern techniques and approaches now aim at determining the mechanisms and roles of the cerebellum as they relate to driving different behaviors in these classic models (15,16). Together, data generated from these different mouse models have cultivated an interest in identifying whether discrete functional features in the cerebellar circuit are the root cause of disease-related behaviors. In this regard, more than half a century since the first descriptions of cerebellar mutants, a core question remains unanswered but hotly debated: how does cerebellar circuit dysfunction lead to unique motor disturbances?

To begin addressing how cerebellar circuits generate behavioral diversity in disease, we used an intersectional genetics approach to mark, map, and manipulate specific types of synapses in the cerebellum. Our approach silences genetically defined populations of synapses by selectively deleting the genes that encode the vesicular GABA transporter, Vgat, or the type 2 vesicular glutamate transporter, Vglut2, from genetically targeted populations of cells. Loss of GABAergic neurotransmission from Purkinje cells, which provide the sole output of the cerebellar cortex, caused uncoordinated movements and disequilibrium that were indicative of ataxia (Figure 1A) (17). In contrast, eliminating glutamatergic neurotransmission from climbing fibers, which synapse onto Purkinje cells, caused twisting postures and hyperextended limbs that is consistent with dystonia (Figure 1A) (18). In addition, systemic injection of the beta-carboline alkaloid drug harmaline caused hyperactivation of climbing fibers and rhythmic bursting spike activity in Purkinje cells as well as a severe tremor (Figure 1A). Optogenetically modulating Purkinje cell to cerebellar nuclei projections in a bursting pattern induced oscillatory tremor movements (19,20). Together, these studies established that different manipulations of Purkinje cell inputs or outputs, and consequently Purkinje cell and nuclei neuron spike signals, can cause diverse behavioral deficits as they relate to disease (18,20,21). Given the heterogeneity and even comorbidity of these behaviors in a single disease model (22), the main question that arises is, do these cerebellar neural signals represent unique pathophysiological signatures that can drive the predominant behavioral defects used to characterize different motor diseases? Here, we aim to resolve this question to provide insight into the origin of cerebellar movement disorder presentation.

Supervised classifier model predicts mouse phenotype based on spike signatures.

A. Schematic of network changes causing motor impairments in mouse models for ataxia, dystonia, and tremor. Dotted lines indicate lack of neurotransmission. Color-coded lines indicate primary affected cell type. B. Example of spike firing rate averaged over previous 50 ms at each occurring spike for the 5 s spike train in B’. B’’ 1 s spike train for the duration indicated in the square box in B’. C. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train used in the classification model. We indicate the firing rate (spikes/s), median ISI-1, skewness, and ISI>25 for all for example cells. D. Classifier model based on training data set (control: n = 25 cells; ataxia: n = 20; dystonia: n = 20; tremor: n = 20). E. Assigned spike signature based on spike properties in complete data set (control: n = 33 cells, N = 9 mice; ataxia: n = 24, N = 5; dystonia: n = 24, N = 9; tremor: n = 24, N = 6). Categories on X-axis indicate the origin of the recorded neurons. F. and G. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons. PC = Purkinje cell; CN = cerebellar nuclei; IO/CF = inferior olive/climbing fiber.

Results

Spike signatures are different between archetypal ataxia, dystonia, and tremor models

We first compared the spike train activity between mouse models for ataxia (L7Cre;Vgatfl/fl, (17,20)), dystonia (Ptf1aCre/+;Vglut2fl/fl, (18,23)), and tremor (harmaline injection, (20,24)) (Figure 1A). We identified these three mouse models as archetypal representations of their respective cerebellar motor disease phenotypes because their overt behavioral motor abnormalities are caused by cerebellum-specific manipulations that do not cause changes in the gross anatomy, cell morphology, or cell survival of the adult cerebellum. These models also exhibit remarkably reliable, severe, and penetrant ataxic, dystonic, and tremor symptoms within each respective group. Video 1 provides examples of the varied and overt motor abnormalities of the mouse models used. Supplemental table 1 summarizes the reported behavioral impairments from prior characterization of these archetypal mouse models.

We hypothesized that these differences in behavioral abnormalities were accompanied by specific changes in the spike train patterns in the cerebellum. We analyzed in vivo electrophysiology recordings of the spike activity in cerebellar nuclei neurons of the interposed nucleus (Figure 1B,C), in awake, head-fixed mice with overt motor phenotypes and control animals. We focused our recordings on the interposed nucleus based on the hypothesis that the cerebellum communicates ongoing motor signals to other brain regions through this nucleus (25) and deep brain stimulation (DBS) of this region successfully reduces motor impairments in mouse models of ataxia (26), dystonia (18), and tremor (20). We described the spike train firing features using twelve parameters that summarized the spike train rate (Figure 1 – Supp. Figure 1A-C), irregularity (Figure 1 – Supp. Figure 1D-F), pauses (Figure 1 – Supp. Figure 1G-I), and rhythmicity (Figure 1 – Supp. Figure 1J-L). In agreement with our hypothesis, we found a significant difference between at least two archetypal groups for each of the twelve spike train parameters. However, none of the twelve parameters showed a statistical difference between all four groups (control, ataxia, dystonia, and tremor), suggesting that the difference between spike train signatures relied on a combination of multiple spike train parameters.

Even though each parameter describes a different feature of the spike train activity, there is a strong correlation between many parameter values in each group of recordings (Figure 1 – Supp. Figure 2). We therefore set out to find the minimal set of parameters, and specific threshold values, that could differentiate the spike train parameters between the archetypal groups. We trained twelve supervised classification learning models based on the spike train parameters of each mouse model (Figure 1 – Supp. Figure 1A-L) with different sets of cells used for training and validation (Figure 1 – Supp. Figure 3). We sought to determine which classifier model best assigned the validation neurons’ spike signatures to the correct origin mouse model. We found that the two classifier models (model 10 and model 9) with the best performance in the validation set used the same parameters and parameter cut-offs to classify the spike signatures between control, ataxia, dystonia, and tremor mice (Figure 1 – Supp. Figure 3). We therefore used this classifier throughout the remainder of the paper to assign a phenotypic signature to the spike train properties of cerebellar nuclei neurons.

The classifier model identified “skewness” (Figure 1D) as the best differentiator between neurons recorded in ataxic and control mice (lower skewness) and neurons recorded in dystonic and tremoring mice (higher skewness). Skewness is a measure of the relative skew in the inter spike interval (ISI) distribution of single neurons from the median to the mean (Figure 1 – Supp. Figure 1F) and is unusually elevated in our dystonic and tremor mouse models (Figure 1C). Next, the classifier model identified CV2 (27) (Figure 1 – Supp. Figure 1E) as the best differentiator of neurons recorded in ataxic mice (lower CV2) from control mice (higher CV2). CV2 measures the irregularity of interspike intervals with less influence from the overall firing rate. It is extraordinarily low in the highly regular spike activity observed in our ataxia model (Figure 1B’’). Last, the classifier model identified that the ISI>25 (Figure 1 – Supp. Figure 1C), which measures the percentage of ISI with a duration over 25 ms (or instantaneous firing rate under 40 Hz) (Figure 1 – Supp. Figure 1G), as the best differentiator of neurons from the mice with tremor (lower ISI>25) compared to those recorded in dystonic mice (higher ISI>25) (Figure 1C). These findings suggest that changes in a combination of three parameters (skewness, CV2, and ISI>25) can differentiate neural spike signatures in mice with distinct behavioral presentation of cerebellar disease.

We assessed whether this classifier model correctly assigns the spike signature corresponding to the behavioral phenotypes of the archetypal mouse models based solely on the spiking activity of single cerebellar nuclei neurons. We tested this by assigning a signature based on skewness, CV2, and ISI>25 values. We found that the classifier model indeed assigned the signature that aligned with different predominant phenotypes of each archetypal mouse model correctly for the majority of neurons (control: 85% of neurons recorded in a control mouse had a control signature; ataxia: 79%; dystonia: 88%; tremor: 54%) (Figure 1E). Our analyses also showed that most neurons of a specific signature were recorded from the mouse model with that phenotype (control: 63% of neurons with a control signature were recorded in a control mouse; ataxia: 93%; dystonia: 74%; tremor: 88%) (Figure 1E-G). Furthermore, the relative proportion of neurons with a control signature was significantly enriched in the control population relative to the entire population (chi-square: p<0.0001) (Figure 1E), and we also found this enrichment for the ataxia signature in the ataxia neurons (p<0.0001), dystonia signature in the dystonia neurons (p<0.0001), and tremor signature in the tremor neurons (p<0.0001). These data support the hypothesis that spike signatures are reliably different between archetypal mouse models of distinct motor behaviors that mimic human disease symptoms.

Multiple spike train signatures can lead to ataxic motor impairments

Next, we set out to identify whether the spike train signatures for the archetypal mouse models are shared across mouse models with different manipulations but similar phenotypes. We started by classifying the spike train parameters in a mouse model of ataxia caused by a poly-glutamate expansion in the Atxn1 gene, Atxn1154Q/+ mice, that causes spinocerebellar ataxia type 1 (SCA1). We measured the spike train patterns early in the disease progression (28) (Figure 2A-B). Despite the presence of an ataxic phenotype at this age (Supp. Table 1), we observed that most Atxn1154Q/+ neurons exhibited a control signature (22/34) and that the second most numerous spike train signature was a dystonia signature (10/24).

Spike signatures in ataxia models with different etiologies.

A 5 s example spike trains (and 1 s inset) of representative cerebellar nuclei neurons recorded in each mutant mouse model. Left: example cell from early disease progression (4-month-old) ataxic Atxn1154Q/+ mouse. Right: example cell from late disease progression (11-month-old) ataxic L7Cre;Ank1fl/fl mouse. B. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train of the example cells in A. We indicate the firing rate (spikes/s), median ISI-1, skewness, and ISI>25. C. Proportion of predicted spike signatures in each of the mouse models. D. and E. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons. Control: n= 27 cells, N = 5 mice; Atxn1154Q/+: n = 34, N = 4; L7Cre;Ank1fl/fl: n = 14, N = 3. C-E. are based on classifier model in Figure 1D.

We reasoned that the high proportion of cells exhibiting a control signature in Atxn1154Q/+ mice might be because the disease-causing genetic abnormality affects the whole body and, consequently, the ataxic symptoms may result from dysfunction in multiple nodes in the motor circuit. We therefore also investigated a mouse model with a cerebellum-specific loss of the gene Ank1 (10-12 month old L7Cre;Ank1fl/fl mice, Figure 2A), which causes an adult-onset degenerative ataxia (29). In this mouse model, we found that half of the neurons exhibited an abnormal spike train signature, with the majority classified with the dystonia spike train signature (6/14).

Our classifier model differentiates the control and dystonia signatures from each other based on the skewness parameter. Previous studies have suggested that the difference between ataxia and dystonia symptoms may be caused by the relative difference in ISI regularity (15,30,31), which can be measured by our skewness parameter. Indeed, the population distribution of skewness values was lower in the ataxic Atxn1154Q/+ (0.14 ± 0.02 (mean ± SEM)) and L7Cre;Ank1fl/fl (0.17 ± 0.03) mouse models (Figure 2D-E) than in our archetypal dystonia model (0.42 ± 0.05) (Ptf1aCre/+;Vglut2fl/fl, Figure 1F-G). We used a one-way ANOVA followed by a Tukey post-hoc analyses and found that the difference in skewness between the ataxia models and the dystonia models was statistically significant (Atxn1154Q/+ vs. Ptf1aCre/+;Vglut2fl/fl: p<0.0001; L7Cre;Ank1fl/fl vs. Ptf1aCre/+;Vglut2fl/fl: p=0.0025), but we did not observe a difference in skewness between the two ataxia models (Atxn1154Q/+ vs L7Cre;Ank1fl/fl: p=0.9961). Together, these findings suggest that there are multiple cerebellar spike train signatures that can lead to ataxic symptoms that both differ from the dystonia signature in their relative level of regularity; unusually regular spike train patterns are seen in L7Cre;Vgatfl/fl mice (Figure 1) and intermediate irregular spike train patterns are observed in Atxn1154Q/+ and L7Cre;Ank1fl/fl mouse models (Figure 2).

Dystonic spike train signatures are shared across etiologically distinct dystonia models

Next, we set out to identify whether the dystonia signature was shared across mice with overt dystonic phenotypes. We included two dystonia models in our analyses; Pdx1Cre;Vglut2fl/fl mice exhibit mild dystonic symptoms only in response to a stressful event (32,33) (Figure 3A-B), whereas mice that receive ouabain applied to the surface of the cerebellum exhibit continuous severe dystonic features (34,35) (Figure 3A-B). We found that the proportion of neurons that exhibited the control signature was greater in Pdx1Cre;Vglut2fl/fl mice than in ouabain infusion mice (7/23 vs. 13/21, z=-2.094, p=0.0366) (Figure 3C). Moreover, the proportion of neurons with a dystonia signature was lower in Pdx1Cre;Vglut2fl/fl mice than in ouabain infusion mice (15/23 vs. 7/21, z=2.1128, p=0.0349) (Figure 3C). Therefore, the mice with a more severe dystonia phenotype had a smaller proportion of cells matching the control signature and a greater proportion of cells matching the dystonia signature. In addition to the proportional differences in cells with a control and dystonia signature, we found that the distribution of skewness values differed between the Pdx1Cre;Vglut2fl/fl (0.12 ± 0.05) and ouabain infusion mice (0.36 ± 0.8) (Figure 3D-E). We quantified this difference by comparing these two mouse models to each other and the archetypal dystonic model (Ptf1aCre/+;Vglut2fl/fl) using a one-way ANOVA followed by a Tukey post-hoc analysis. We found that the skewness was lower in Pdx1Cre;Vglut2fl/fl mice with mild dystonic features compared to the ouabain infusion mice (p=0.0012) and Ptf1aCre/+;Vglut2fl/fl mice (p<0.0001) with severe dystonic features. There was no significant difference between the two models with severe dystonic features (p=0.9274). These findings show that the dystonia signature is shared across etiologically distinct dystonia models and that the relative severity of symptoms may be explained by the proportion of neurons with a dystonia and control signature (Figure 3).

Spike signatures in dystonia models with different etiologies.

A 5 s example spike trains (and 1 s inset) of representative cerebellar nuclei neurons recorded in each mutant mouse model. Left: example cell from mildly dystonic Pdx1Cre;Vglut2fl/flmouse. Right: example cell from severely dystonic cerebellum ouabain infusion mouse. B. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train of the example cells in A. We indicate the firing rate (spikes/s), median ISI-1, skewness, and ISI>25. C. Proportion of predicted spike signatures in each of the mouse models. D. and E. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons. Control: n= 28 cells, N = 6 mice; Pdx1Cre;Vglut2fl/fl: n = 23, N = 5; cerebellum ouabain infusion: n = 21, N = 4. C-E. are based on classifier model in Figure 1D.

Tremor spike train signatures are shared across etiologically distinct tremor models

Next, we set out to identify whether the tremor signature is shared across mice that exhibit severe oscillatory tremors. To test this, we used Car8wdl/wdl mice. We previously showed that Car8wdl/wdl mice exhibit a severe oscillatory tremor that can be reduced to control levels by treatment with the beta blocker drug propranolol (36), which is used to treat tremor in human patients. Car8wdl/wdl mice also exhibit ataxia and dystonia, which are not normalized by propranolol (10,22,26,36). We found that, before treatment, Car8wdl/wdl mice exhibited a large proportion of neurons with the tremor signature (8/13). However, in propranolol-treated Car8wdl/wdl mice the proportion of neurons with the tremor signature was diminished while control (6/15) and dystonia (6/15) signatures were the most common signatures (Figure 4C). The classifier model differentiates between the dystonia and tremor signature based on the ISI>25 value and, indeed, we find a shift towards higher ISI>25 values in propranolol-treated Car8wdl/wdl mice compared to untreated Car8wdl/wdl mice (Figure 4D-E). Together, these findings suggest that the tremor signature is shared across etiologically distinct mouse models of tremor and that the tremor-reducing drug propranolol also reduces specific spike train features contributing to the tremor signature (Figure 4).

Spike signatures in tremor models with different etiologies.

A 5 s example spike trains (and 1 s inset) of representative cerebellar nuclei neurons recorded in each mutant mouse model. Left: example cell from Car8wdl/wdlmouse with complex phenotype including severe tremor. Right: example cell from Car8wdl/wdl mouse with complex phenotype treated with propranolol to treat tremor. B. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train of the example cells in A. We indicate the firing rate (spikes/s), median ISI-1, skewness, and ISI>25. C. Proportion of predicted spike signatures in each of the mouse models. D. and E. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons. Control: n= 14 cells, N = 4 mice; Car8wdl/wdl: n = 13, N = 6; Car8wdl/wdl + propranolol: n = 15, N = 6. C-E. are based on classifier model in Figure 1D.

Different spike signatures can be generated by the same canonical cerebellar circuit

Thus far, the observed spike train signatures have been dependent on developmental changes in circuit connectivity (L7Cre;Vgatfl/fl; Ptf1aCre/+;Vglut2fl/fl; Pdx1Cre;Vglut2fl/fl), neurodegeneration (Atxn1154Q/+; L7Cre;Ank1fl/fl), the effects of transgene expression (Atxn1154Q/+) or frameshifting (Car8wdl/wdl), or drug effects on the nervous system outside of the cerebellum (harmaline, ouabain, propranolol). Therefore, to determine whether the distinctive features of spike activity could be generated by the same, otherwise healthy circuit, we expressed a light-sensitive cation channel opsin, channelrhodopsin (ChR2), in Purkinje cells (L7Cre;ROSA26loxP-STOP-loxP-EYFP-ChR2 mice, hereon referred to as L7Cre;R26ChR2 mice). Upon light activation, Purkinje cell firing rate increases (Figure 5 – Supp. Figure 1) and nuclei neurons are inhibited via the GABAergic neurotransmission from Purkinje cells (Figure 5A-C). We leveraged this circuit connectivity and optogenetic strategy to stimulate Purkinje cell terminals in the interposed nucleus to induce distinct spike signatures in those downstream interposed nucleus neurons (Figure 5). We chose three stimulation paradigms. The ataxia stimulation was a continuous 50 Hz square pulse stimulation (mimicking the lack of modulation by Purkinje cells in our degenerative and genetically silenced ataxia models); the dystonia stimulation was a 50 Hz square pulse stimulation that was on for at least 1 s with a 75% chance of a 250 ms pause between periods of stimulation (to induce the slow and irregular spike train features of our defined dystonia signature); and the tremor stimulation was a 10 Hz sinusoidal stimulation (to induce the fast, irregular, and rhythmic spike train features of our defined tremor signature). We performed in vivo recordings in awake, head-fixed mice, and confirmed that each of the optogenetic stimulation paradigms caused distinct changes in Purkinje cell spiking activity (Figure 5 – Supp. Figure 1). We then directed stimulation towards Purkinje cell terminals surrounding interposed nucleus neurons. The stimulation paradigms induced distinct changes in the spike train features of interposed nuclei neurons using this method (Figure 5D). We then compared the resultant spike train features to the defined thresholds of our classifier model. We found that the ataxia stimulation paradigm induced cells to develop a response similar to the Atxn1154Q/+ and L7Cre;Ank1fl/fl ataxia model mice (Figure 3, Figure 5 – Supp. Figure 2), with about half of the cells remaining within the control signature (7/16), and half shifting into the dystonia signature (8/16) (Figure 5F). Also akin to these models, when we used a one-way ANOVA followed by a Tukey post-hoc analysis to compare the ataxia and dystonia stimulation responses, we found the ataxia stimulation resulted in significantly lower skewness than the dystonia stimulation (ataxia stimulation versus dystonia stimulation: p=0.0287). Additionally, we found that the dystonia and tremor stimulation induced the majority of cells to produce spike signatures that our classifier model defined as dystonic (9/16 cells) or as tremor signatures (11/16 cells), respectively (Figure 5F). These data indicate that spike signatures associated with ataxia, dystonia, and tremor can be generated by the same, otherwise healthy, cerebellar circuit.

Spike signatures in cerebellar nuclei neurons can be induced by specific stimulation paradigms of Purkinje cells.

A. Schematic of experimental setup with recordings in awake, head-fixed mice. B. Optopatcher recordings of cerebellar nuclei neurons. The opsin is expressed in Purkinje cells (pink) and recordings of nuclei neurons are performed. PC = Purkinje cell; CN = cerebellar nuclei; IO/CF = inferior olive/climbing fiber. C. Example validation that light stimulation of inhibitory Purkinje cells (blue bars) inhibits nuclei neurons during light stimulation. The lower trace is a blown-up view of the boxed area in the upper trace. D. Example of spike firing rate averaged over previous 50 ms at each occurring spike for the 5 s spike train in D’. D’’ 1 s spike train for the duration indicated in the square box in D’. Blue bars indicate light stimulation and are specific for ataxia, dystonia, and tremor (see methods for light stimulation parameters). All example traces originate from the same nuclei neuron, indicating that the cell’s spike signature can change depending on the light stimulation paradigm. E. Histograms of instantaneous firing rate (ISI-1) of the full 30 s spike train, observe the shift in distribution from baseline during the different stimulation paradigms. F. Proportion of cells of each predicted spike signature during each of the light stimulations based on classifier model in Figure 1D. Control: n = 26 cells, N = 7 mice; ataxia: n = 16, N = 6; dystonia: n = 16, N = 5; tremor n = 16, N = 5. G. and H. Scatterplot of spike train parameters used to classify neural signatures. Colored boxes indicate the predicted phenotype, colors of circles indicate the origin of the recorded neurons.

Spike signatures can induce distinct motor phenotypes that mimic disease-related behaviors

During the head-fixed stimulation experiments, we observed subtle behavioral responses during unilateral cell-targeted optogenetic stimulation that suggested the different spike signatures might drive unique motor behaviors (Video 2). To explore this further in freely moving mice, we investigated whether the different spike signatures directly drive unique motor disturbances when a population of cells are induced to produce our defined spike signatures. To this end, L7Cre;R26ChR2 mice were implanted with optical fibers that were bilaterally targeted to the interposed cerebellar nuclei (Figure 6A-C). This allowed for the targeting of Purkinje cell terminals in the cerebellar nuclei, as was done during our in vivo recordings in Figure 5, but with a larger population of cells affected by the stimulation. This also allowed the mice to move freely during stimulation, making a series of behavioral assays during stimulation possible (Figure 6D). Changes in behavior were immediate and severe as soon as the stimulation began. Each stimulation paradigm resulted in a different constellation of behaviors with the ataxia stimulation paradigm resulting in imprecise movements, the dystonia stimulation paradigm resulting in prominent dystonic posturing, and the tremor stimulation paradigm resulting in severe tremor (Video 3). Therefore, we assessed the mice for changes in gait, presence of dystonic movements, and severity of tremor. No single measurement can perfectly encapsulate the complexity of ataxia, dystonia, and tremor – all of which can appear in combination with the others (37–39). Additionally, each phenotype may affect the measurement of others. This was particularly apparent with measurements of gait. We found that all stimulation paradigms affected gait, resulting in visibly different foot placement (Figure 6E-F) and movement down the footprinting corridor (Video 4). Quantitatively, this was evident in shorter steps and less precision of fore and hindpaw placement (Figure 6I-K). We also considered the behavior of the mice in an open, flat, plexiglass arena for signs of dystonic movements (Figure 6L, Video 3). The dystonia stimulation paradigm resulted in the most frequent and strongest dystonic movements, while we also observed dystonic movements with the ataxia paradigm and abnormal movement – namely severe tremor – was noted with the tremor stimulation paradigm (Figure 6M). Mice were also placed in a custom-made tremor monitor (20) where they could freely ambulate during stimulation while an accelerometer mounted under the arena detected changes in acceleration (Figure 6D). While the ataxia stimulation paradigm did not significantly increase tremor from baseline, both the dystonia and tremor stimulation paradigms resulted in a significantly increased tremor at 10 Hz (Figure 6N-P). Though dystonia is often observed with tremor, it is possible that the detection of tremor in these animals was due to the jerkiness of their dystonic movements rather than an increase in smooth, rhythmic tremor that is more often associated with tremor disorders. Indeed, the tremor paradigm resulted in the most severe tremor of all stimulation conditions, producing a tremor severity that was more than ten-fold greater than that of the dystonia stimulation paradigm. Together, these measurements produced a complex representation of the underlying phenotypes. All three stimulation parameters resulted in a behavioral repertoire that was significantly different from baseline and, while there was some overlap of features, each stimulation paradigm produced the distinct and predicted respective motor phenotype (Figure 6Q-T). Together these data suggest that our classifier model’s defined spike signatures of disease-associated cerebellar nuclei spike trains are sufficient to produce the predicted corresponding abnormal motor phenotypes in mice.

Induced spike signatures elicit distinct cerebellar phenotypes.

A. Schematic of external view of bilateral optical fiber implant. B. Schematic of a coronal section from a mouse cerebellum with bilateral optic fiber implants directed towards the cerebellar nuclei. FN = fastigial nucleus. IN = interposed nucleus. DN = dentate nucleus. C. Photomicrograph of a Nissl-stained coronal section from a mouse cerebellum that had been implanted with optic fibers. Arrows = optic fiber tracks. Dotted lines surround the cerebellar nuclei indicated in B. Scale = 1 cm. D. Schematic of a mouse with bilateral optic fiber implants freely moving in a tremor monitor. E-K. Data associated with gait measurements. E-H. Example footprints from a single mouse before stimulation (baseline, E) and during ataxia (F), dystonia (G), and tremor stimulation (H). Scale = 1 cm. I-K. Measurements of gait including the length of the hindpaw stride (I), forepaw stride (J), and distance between the hind and forepaws (K). N = 8 mice. * = p ≤ 0.05. ** = p ≤ 0.01. L. Example images of phenotypes associated with dystonia. 1 = erect tail. 2 = high stepping. 3 = kinked tail. 4 = hyperextension of the limbs. 5 = splayed toes. M. Dystonia rating of mice before stimulation and during stimulation with each paradigm. N = 8 mice. * = p ≤ 0.05. N. Tremor signals detected via tremor monitor from a mouse before and during stimulation with each paradigm. Horizontal scale = 1 sec. Vertical scale = 50 mV. O. Population average power spectrums of tremor. Solid line = mean power. Shaded region = SEM. P. Peak tremor power of mice before and during stimulation with each paradigm. N = 8 mice. * = p ≤ 0.05. ** = p ≤ 0.01. Q-S. 2-D comparisons of gait (hind to fore distance), dystonia (rating), and tremor (peak power) measurements from all mice. N = 8 mice. T. 3-D plot of data in Q-S.

Discussion

In this study, we tested whether distinct spike train signatures in the interposed cerebellar nuclei explain why cerebellar dysfunction can cause multiple distinct motor impairments associated with movement disorders. By comparing spike activity across multiple mouse models of cerebellar disease, we found that the cerebellum can generate a range of dysfunctional spiking patterns. We found disease-specific spike train signatures using a classifier learner model, which allowed us to discover specific spike train parameters and their corresponding cutoff values that could distinguish the activity associated with these different disease states. When investigating whether these spike signatures are shared across mouse models with similar phenotypes due to different etiologies, we found that two types of spike train activity can cause ataxia, whereas specific spike train signatures are strongly associated with dystonia and tremor. We then tested whether we could optogenetically induce these signatures in an otherwise normal circuit. We found that the same neurons can generate healthy spike activity as well as a spectrum of disease-like spike activities. We then tested whether these optogenetically-induced disease-associated spike signatures could elicit similar behavioral abnormalities in the absence of any other primary genetic or circuit defects in awake and freely moving mice. The predominant behaviors that characterize each disease condition were successfully recapitulated. These data provide compelling evidence for the reliance of neurological phenotype presentation on the pattern of cerebellar circuit misfiring.

Several previous studies have proposed that distinct spike train patterns may correspond to distinct presentations of cerebellar disease (15,30,31,40). Our work builds on these studies by quantitatively comparing spike train properties across, rather than within, mouse models. We demonstrate which aspects of the spike train patterns are distinct and shared across mouse models with different and similar disease phenotypes, respectively. We confirm prior research indicating the importance of spike train irregularity in disease presentation (15,30,40). We also show that spike train irregularity is insufficient to differentiate the spike train properties of dystonic and tremoring mice, which are differentiated from each other based on the instantaneous firing rate rather than pattern. Additionally, we confirm that the spike train signatures associated with different disorders can cause distinct disease phenotypes, thereby showing for the first time that distinct cerebellar spike train signatures are sufficient to drive a variety of motor impairments. Together, these data provide strong evidence that different spike train signatures do not only result from sensory feedback towards the cerebellum and that they may be a primary cause for motor impairments associated with cerebellar disease.

It is intriguing that we find multiple cerebellar spike signatures associated with an ataxic phenotype. Our work associates L7Cre;Ank1fl/fl and Atxn1154Q/+ ataxias with slightly irregular (elevated skewness) and slow or pause-prone (ISI>25) firing patterns (Figure 2). However, the L7Cre;Vgatfl/fl ataxia is associated with very regular firing patterns on the scale of both spike-to-spike irregularity (CV2) and that of the entire trace (skewness) (Figure 1). In this case, we feel it is essential to consider the underlying pathogenesis of these mouse models (Table 1). The L7Cre;Vgatfl/fl mouse represents a non-degenerative ataxia with a cell type-specific cerebellar insult (17). However, while L7Cre;Ank1fl/fl is cerebellar cell type-specific, it is also neurodegenerative (29). Meanwhile, Atxn1154Q/+ is neither cell type nor cerebellar-specific and is neurodegenerative (41). Therefore, our work suggests that degenerative ataxias and non-degenerative ataxias may have different underlying circuit alterations that lead to distinct spike signatures. We also consider the possibility that the complete lack of Purkinje cell GABA neurotransmission in the L7Cre;Vgatfl/fl mouse allows this high degree of regularity in the cerebellar nuclei cells, while the L7Cre;Ank1fl/fl and Atxn1154Q/+ mutations likely alter the input from Purkinje cells that the nuclei cells receive, but do not completely remove it. The possibility of any modulation from Purkinje cells likely enables, and perhaps ensures, greater irregularity than that which is possible in the L7Cre;Vgatfl/fl model. This is somewhat supported by our awake, head-fixed recordings of nuclei cells in response to Purkinje cell terminal channelrhodopsin stimulation, in which we were able to achieve firing patterns more similar to the L7Cre;Ank1fl/fl and Atxn1154Q/+ signatures than that of L7Cre;Vgatfl/fl (Figure 5 – Supplemental Figure 2). Our channelrhodopsin experiment attempts to minimize the modulation of the Purkinje cell input to the cerebellar nuclei, but there is no more minimal modulation of an input than the complete lack of it. However, an important caveat to this interpretation is that our manipulation of the cerebellar nuclei during the awake recordings only impacted Purkinje cell terminals near the nuclei cell being recorded, by design. Therefore, it is possible that with greater synchrony of Purkinje cells (I.e., mimicking the pan-Purkinje cell silencing in the L7Cre;Vgatfl/fl mouse) a more regular nuclei firing pattern could be induced, if the Purkinje cell activity itself has a very regular firing pattern (42). Indeed, something closer to this may have been achieved with our bilateral implanted optic fibers with our behavioral experiments. These experiments resulted in distinct behavioral outcomes for each stimulation pattern (Figure 6). In short, we have uncovered an intriguing divergence in the spike signatures that are associated with an ataxia-like phenotype. We anticipate that studies into cerebellar neurodegeneration and Purkinje cell synchrony in the context of reduced dynamic range may be fruitful in exploring how these two cerebellar signatures may impact ataxic phenotypes.

Our work suggests that there is a healthy range within the characteristics of cerebellar nuclei spiking activity and that cerebellar movement disorders are associated with a shift from this range in one or multiple features of spike train activity. We find some overlap and shared spike features between the different disease phenotypes and show that healthy cerebellar neurons can adapt multiple disease-associated spike train signatures. These findings suggest that pathophysiological spike train signatures are driven by a shift in neural function towards a disease state that can be independent of plasticity or circuit rewiring. Despite the dramatically different behavioral outcomes, the potential overlap and shared spike features, such as the irregular spike pattern found in the dystonia and tremor signatures, raise the strong possibility that co-expression or comorbidity of different motor disease behaviors may arise due to a spectrum of spike signal defects. This indicates that it is possible for neural signals to shift back and forth between healthy and disease states, potentially resulting in the transientness of behavioral impairments in certain cerebellar disorders. These findings also suggest that the most successful therapeutic avenues for cerebellar movement disorders should maintain cerebellar spiking activity within the healthy range without inadvertently inducing a different pathophysiological signature.

One important question to consider is the identity of the cells that produce these disease-relevant spike signatures. Both Purkinje cells and cerebellar nuclei cells are heterogeneous populations with varied electrophysiological properties (43–46). The recordings we report here were extracellular and performed in awake, head-fixed mice and, therefore, we cannot speak to more precise identities of the cell population included in this work. We can speculate that, as excitatory nuclei cells are larger than inhibitory nuclei cells (45), our recordings of nuclei cells are more likely to be from excitatory neurons. However, our optogenetic technique uses Purkinje cell terminals to modulate nuclei activity and therefore would impact both excitatory and inhibitory nuclei neurons (47). Recent work has suggested differential therapeutic benefit and accessibility for treatment for excitatory and inhibitory neurons (48). Therefore, future studies parsing the impact of excitatory and inhibitory cerebellar neurons on motor impairments would be a great benefit to the field.

There has been (49), and there still remains (50), a great interest in understanding how altered cerebellar signals influence human behavior and disease. A pressing need to better define the architecture of these spiking abnomalities has arisen because of the success in using deep brain stimulation (DBS) and the potential for better tuning of the stimulation parameters for greater efficacy in treating different cerebellar-related disorders (ataxia: (51,52); dystonia: (53); tremor: (54,55); ataxia, dystonia, tremor: (50)). Recordings in human patients during DBS implantation have previously found physiological differences in the spike train patterns of basal ganglia neurons between dystonia and other movement disorders (for example; etiologically distinct dystonia: (56); dystonia with and without head tremor: (57); dystonia versus Parkinson’s disease: (58)). These works and ours compaired transient, opportunistic recordings of neurons and therefore contain variability imparted from unknown levels of arousal or sensory states. Here we show that these nonsimultaneous recordings from a population of heterogeneous neurons is suffient to determine distinct pathogenic signatures amongst the variability of our dataset. However, simultaneous recordings may reveal greater insights into the neural signatures of motor diseases. Simultaneous recording may be beneficial across nodes in the motor circuit (for example, the basal ganglia and cerebellum in dystonia) or to consider synchrony within a cell population (for example, across Purkinje cells in ataxia). Indeed, simultaneous recordings can be valuable in determining upstream and downstream nodes in the propagation of pathogenic neural activity, especially if paired with the appearance of disordered movements (59), which is useful in determining potential therapeutic targets. Our electrophysiological data confirm that differences between disease presentations are found in the spike train properties of the cerebellum and our optogenetic experiments suggest that these spike trains are not merely biomarkers that correlate with the disease state but are sufficient to cause motor impairments. Therefore, it is not clear whether our model would become stronger by considering both cerebellar activity and activity elsewhere in the motor circuit or if considering multiple areas would be redundant. Our previous work and what we have shown here suggests cerebellar-targeted DBS may be highly beneficial in normalizing disease-causing spike train patterns. As we move towards greater understanding of the motor system’s response to such perturbation, it would be beneficial to consider the simultaneous responses of neural activity within and external to the cerebellum to determine ideal biomarkers and therapeutic targets to improve DBS for movement disorders.

A parallel motivation has fueled rodent studies to define the anatomical targets, stimulation parameters, and outcome efficiacy in detail (18,26,60–62). Although the exact mechanism(s) of DBS remains unclear (61,63), there is strong support that at least one target of DBS is the actual neuronal signal itself (which the DBS may modify, enhance, dampen, and perhaps even interfere with the given neural spike defects). Indeed, the ability to alter cerebellar spike activity (64) has driven the investigation of the signal properties (49). Evidently, neurotransmission of the faulty activity patterns is required for the expression of disease behaviors (19). Recent works strongly support the hypothesis that changes in activity (rate, pattern, or both) could instigate a range of movement abnormalities (18,20,34,65,66). However, the current work defines the individual potential of these altered cerebellar signals to initiate specific motor changes. The spike signature assocaited with neurodegenerative ataxias (L7Cre;Ank1fl/fl and Atxn1154Q/+) has slightly elevated skewness and slower or pause-prone firing compared to control (ISI>25, Figure 5 – Supplementary Figure 2) while excedingly steady activity is assocaited with the nondegenerative L7Cre;Vgatfl/fl ataxia. Modulation of interposed nucleus activity alters movement trajectories (67). Therefore incorrect modulation of nuclei activity in the neurodegenerative ataxias or lack of correct modulation in L7Cre;Vgatfl/fl ataxia may lead to the ataxia phenotype. The dystonic models herein were characterized with irregular and slow firing. Previous work has associated bursts of Purkinje cell activity with dystonic postures (68) and therefore the suppressed periods of the irregular pattern associated with dystonia may be related to the timing of dystonic muscle contractions. Indeed, in our previous work we have shown that stimulation of Purkinje cell terminals in the cerebellar nuclei, resulting in suppresion of cerebellar nuclei activity, directly precedes muscle contractions (20). Finally, a rhythmic burst pattern of activity was associated with a tremor signature. A large body of work implicates ocillatory activity propagating through the motor network to induce rhythmic oscillations of muscle activity in tremor (20,59,69). Indeed, bursting activity in the cerebellar nuclei has a rather direct relationship to the resultant tremor (20). For all phenotypes, we expect that a large proportion of the population of neurons must produce the abnormal spike signature in order to produce a severe motor phenotype (Figures 2 and 3). Our optogenetic behavioral studies likely induced synchrony of both Purkinje and nuclei cell activity in addition to the intended spike signature, therefore our work here supports synchronous abnormal firing for the phenotypes we produced, however we cannot claim that synchrony is necessary to produce these phenotypes. Together, these data unveil a critical role for the cerebellum in triggering disease behaviors, with cauasative signals likely originating locally in its circuitry. Each disease may arise as a result of a change in the balance and representaion of neural signatures.

Methods

Animals

Mice were housed in a Level 3, AALAS-certified vivarium. The Institutional Animal Care and Use Committee (IACUC) of Baylor College of Medicine (BCM) reviewed and approved all experimental procedures that involved mice. We used the following transgenic mouse lines: L7Cre (also known as Pcp2Cre) (70), Vgatfl (JAX:012897) (71), Ptf1aCre (JAX:023329) (72), Vglut2fl (JAX:012898) (73), Pdx1Cre (JAX:014647) (74), Car8wdl/wdl (JAX:004625) (10), Atxn1154Q (JAX:005601) (41), Ank1fl (JAX:036512) (75), and Ai32 (Rosa26lsl-ChR2-eYFP, JAX:024109) (76). We included mice from both sexes. We used ear punches from pre-weaned pups for PCR genotyping to identify the different transgenic alleles.

Headplate and craniotomy surgery for electrophysiology recordings

Prior to all recordings, we performed a surgery to attach a headplate over bregma and make a craniotomy over the interposed nucleus. This allowed us to collect stable recordings of cerebellar neuron activity while the mouse was awake and with or without a severe motor phenotype. This surgery was detailed in our previous publication (77). In short, mice were given preemptive analgesics including a subcutaneous injection of buprenorphine and meloxicam. Anesthesia was induced using inhaled isoflurane. During the surgery, the mice were kept on a heating blanket to maintain body temperature. Fur was removed from the surgery site using depilatory cream (Nair) and a small incision in the skin was made over the skull. Next, we used a dental drill to make a ∼2 mm diameter craniotomy over the interposed nucleus (6.4 mm posterior and 1.3 mm lateral to Bregma). The craniotomy was surrounded by a round chamber and filled with antibiotic ointment until the day of the recording. The recording chamber was closed with a screw top or silicone cap. We also placed a headplate with a hole over Bregma. A small piece of wire was placed vertically over Bregma to identify this anatomical marker on the day of recording. The recording chamber and headplate were attached to the skull using a combination of C and B Metabond Adhesive Luting Cement (Parkell) and methyl methacrylate dental cement (A-M Systems). Mice were allowed to recover from the surgery for at least three days before we began conducting the in vivo electrophysiological recordings.

Standard in vivo electrophysiological recordings in awake mice

We performed in vivo electrophysiology recordings according to experimental procedures detailed in our previous publications (20,77,78). In brief, we placed mice on a rotating foam wheel and stabilized their heads by screwing the implanted headplates to the recording rig. We removed the antibiotic ointment from the recording chamber and replaced it with sterile physiological saline solution (unless otherwise specified). We measured the coordinates of Bregma to determine the coordinates where the electrode would penetrate the surface of the cerebellum. We used tungsten electrodes (2-8 MΩ, Thomas Recording) attached to a preamplifier head stage (NPI Electronic instruments) for our recordings. The position of the recording electrode was controlled using a motorized micromanipulator (MP-255; Sutter Instrument Co.). Neural signals were amplified and bandpass filtered (0.3-13 kHz) using an ELC-03XS amplifier (NPI Electronics) before being digitized using a CED board. All signals were recorded using Spike2 software (CED). We only included neurons recorded between 2.5 and 3.5 mm from the surface that did not exhibit complex spikes (characteristic for Purkinje cell firing) in our analyses of cerebellar nuclei neuron firing patterns. This paper includes reanalyzed data from previously reported studies (18,20,29,78).

Spike sorting and analyses

We previously showed that recording duration influences the estimation of neural firing parameters (79). Therefore, all parameter estimates in this paper are based on 30 seconds (s)-long recordings. We used Spike2 software to sort out spikes from electrophysiological recordings. Information about spike timing were imported and saved in MATLAB databases using custom written code. We described the spike analyses using twelve parameters investigating the properties of interspike intervals (ISI) within the recording; (1) ; (2) mean instantaneous firing rate = mean (ISI−1); (3) mean instantaneous firing rate = median (ISI−1); (4) ; (5) ; (6) ; (7) ISI = % ISI > 25 ms; (8) ISI100 = % ISI > 100 ms; (9) inter burst pause = mean (ISI > (5 * mean (ISI)); (10) kurtosis = % ISI at mode(ISI); (11) ; (12) oscillation peaks = number of peaks “a”. For the calculation of 11 and 12, we performed an autocorrelation analysis for all spikes in the 30 s recording, calculated the rhythmicity index, and found oscillation peaks as previously described (80,81). We used a bin width of 5 ms. The first oscillation peak was determined as the highest bin between 10 ms and 1.5 times the mean ISI for a given neuron. The difference between the baseline level and the height of the peak was denoted as a1, the difference between baseline and the depth of the trough was denoted as b1, and z was the difference between baseline and the total number of spikes included in the analyses. Each subsequent peak was determined as the highest bin between the delay-time of the previous trough and an+a1+10 ms, where an is the time of the previous peak. The first trough was determined as the lowest bin between the first peak (a1) and an + a1. Peaks and troughs were only accepted if their sum was higher than four times the standard deviation of autocorrelation between 0.96 and 1 s lag-time, or if the peak was higher than baseline plus two times the standard deviation and the trough was lower than baseline minus two times the standard deviation.

Supervised classifier learner

We trained our classifier learner using cerebellar nuclei neuron recordings from mixed background control mice (control), L7Cre;Vgatfl/fl mice (ataxia), Ptf1aCre;Vglut2fl/fl mice (dystonia), and harmaline-injected mice (tremor). We have previously tested these models and consider them as reliable archetypical models for cerebellar movement disorders (17,18,20). For the analyses, we reanalyzed previous recordings (18,20,29,36,78) in addition to newly acquired ones. We previously found that the recording duration used for parameter analyses influences the presentation of spike properties. Therefore, we only included recordings with a duration of exactly 30 s and calculated parameter estimates as described above (79). Our previous study also showed that cerebellar nuclei neurons recorded in the same mouse had equal parameter variability as nuclei neurons recorded in different mice. Therefore, we treated all single nuclei neuron recordings as independent datapoints (79).

We used the built-in supervised machine learning, MATLAB Classification Learner APP (The Mathworks Inc, version R2021a) to define spike signature parameter value cutoffs. We imported the parameters describing the spiking activity from an Excel (Microsoft) worksheet. We used “Group” (control, ataxia, dystonia, or tremor) as the “Response Variable” and the 12 parameters described above as the “Predictor Variables.” We trained a “Coarse Tree” with maximum number of splits = 3, split criterion = Gini’s diversity index, and surrogate decision splits = off. We exported the trained model to the workspace to validate the classifier learner. We chose to use a coarse classifier learner to find unique spike signatures because spike parameters are highly correlated with each other (Figure 1 – Supp. Figure 2) and we wanted to prevent overfitting of our model to correlated parameters based on a relatively limited set of recordings.

We trained twelve classifier models based on training sets of control (n=25), ataxia (n=20), dystonia (n=20), and tremor (n=20) neurons, and twelve distinct validation sets of control (n=8), ataxia (n=4), dystonia (n=4), and tremor (n=4) neurons. We validated how the classifier models (see Figure 1 – Supplemental Figure 3) performed and used the classifier with the best performance throughout the paper to assign signatures to nuclei cell spike patterns.

Optopatcher in vivo electrophysiology recordings in awake mice

Optopatcher experiments were performed as previously described (82). We prepared the mice for recording by performing a headplate and craniotomy surgery as described above. The electrophysiology rig and setup were the same as described above with the following differences: We used an optopatcher device (ALA Scientific Instruments Inc) that allows for the placement of a custom-made optical fiber (Thorlabs, #FT200UMT) within a glass recording electrode (Harvard Apparatus, #30-0057). The tip of the optical fiber was placed near the tip of the recording electrode and was illuminated via a 465 nm LED light source (ALA Scientific Instruments Inc). On the day of the recording, we pulled glass electrodes (Sutter Instruments, #P-1000), filled the electrodes with physiological saline, and measured their impedance using an ELC-03XS amplifier (NPI Electronics) before recording. Only electrodes with 2-15 MΩ impedance were used. Light stimulation was triggered using custom Spike2 scripts paired with a CED board (CED). All optopatcher recordings were performed in L7Cre;Ai32 mice that express channelrhodopsin in Purkinje cells. Nuclei neurons included in our analyses were cell recordings between 2.5-3.5 mm from the surface of the cerebellum and were inhibited by brief light stimulation at 465 nm. After we found and isolated a cell, we slowly ramped up the brightness of this brief stimulation until we found the minimal light power that modulated the spiking activity. This minimal power was then used to stimulate the cell with the various test parameters during the recording session. Only cells with a “control” spike signature at baseline were considered for further analysis.

Optogenetic light stimulation paradigms for optopatcher recordings and behavioral assays

We used the following stimulation parameters to measure signature-specific responses; ataxia: 50 Hz (10 ms light on/ 10 ms light off) square pulses; dystonia: at least 1 s of 50 Hz (10 ms light on/ 10 ms light off) square pulses interspersed with at 75% chance of 250 ms pauses in stimulation; tremor: 100 Hz pulses (5 ms light on/ 5 ms light off) overlaid with a 10 Hz sinusoid (50 ms parabolic increase and decrease of light power followed by 50 ms light off).

Bilateral optic fiber implant surgery for in vivo behavioral assays

We implanted mice with optical fibers (ThorLabs, #FT200UMT) targeted bilaterally to the interposed cerebellar nuclei to assess the motor phenotypes that result from our stimulation parameters. We previously described this surgical procedure (Brown et al., 2020). Briefly, the mice were given subcutaneous injections of sustained-release buprenorphine and meloxicam as preemptive analgesics. Anesthesia was induced with 3% isoflurane gas and the anesthetic plane was maintained with 2% isoflurane gas. The mice were placed on a heating blanket and the head was stabilized in a stereotaxic frame (David Kopf Instruments). Fur was removed from the top of the head using depilatory cream (Nair). The surgical site was cleaned using alternating applications of betadine scrub and alcohol. An incision was made in the skin to expose the skull from anterior to bregma to the posterior aspect of the occipital plate. The fascia was removed from the top of the skull and a scalpel was used to etch fine grooves into the top of the skull. A small craniotomy was made in the parietal plate with a dental drill in order to attach a skull screw (00-90 x 1/16 flat point stainless steel machine screw) to anchor the implant to the skull. The skull screw was advanced only until the point that it was stable in the skull and care was taken to ensure it did not contact the brain. The implant sites were determined by measuring 6.4 mm posterior and ±1.5 mm lateral to bregma. Small craniotomies were made at these sites using a dental drill. The base of the fibers were placed on the surface of the cerebellum and were advanced ventrally for 2 mm. A small amount of antibiotic ointment was placed around each fiber to prevent Metabond or dental cement from entering the craniotomy. Metabond was applied over the entire exposed area of skull and around the skull screw and fibers. Two short segments of wire were embedded in the Metabond to allow the experimenter to hold the mouse’s head while attaching and removing the fiber patch cables from the implant. Dental cement was placed over the Metabond. The mice were allowed to recover for at least 3 days before any behavioral assessments were made. Subcutaneous injections of meloxicam were provided every 24 h during the recovery period. Eight mice (number of mice, N) were implanted and tested with all three behavioral assays.

Gait analysis

Measurements of gait were made for all mice both before (baseline) and during each stimulation parameter. To do this, the mice were briefly scruffed and then contrasting colored non-toxic paints were applied to the soles of their forepaws and hindpaws (blue paint for forepaws and red paint for hindpaws). The mice were then gently set down on a piece of blank white paper positioned between two parallel plexiglass barriers with a dark enclosed area at the end of the corridor. Adult mice are naturally inclined to walk towards the dark enclosed area, meanwhile depositing paint on the paper with each footstep. The stimulation, if present, was extinguished once the mice entered the enclosed area. The mice were allowed to rest for at least 2 min between each trial. A trial was considered successful if there were at least four hind and forepaw prints for both the left and right feet that were visible, steady (the mouse was neither running or stopping midway through the series of footsteps), and in as straight of a line as possible (the mouse was not actively turning during the trial). Optic patch cables were connected to the implant during every trial (including baseline runs when stimulation was not present). At least two successful trails were collected per stimulation parameter. For analysis, measurements from two trials of the same type were averaged to determine the values for each mouse. These measurements were stride, the distance from one footprint to the next from the same foot and hind to fore distance, and hind-to-fore distance, the distance from one hindpaw placement from a corresponding forepaw placement on the same side of the body. Three of each measurement were made per footprint and were averaged to determine the measurements per foot, per trial. These trials were then averaged to determine the final measurements for each mouse. A repeated-measures one-way ANOVA with a Tukey multiple comparison adjustment was performed to determine significance. Significance values were indicated as not significant (ns) if p > 0.05, * = p ≤ 0.05. ** = p ≤ 0.01.

Dystonia rating scale

Mice were placed in a rectangular plexiglass arena in order to phenotypically rate the frequency and severity of dystonic behavior as described previously (Jinnah et al., 2000; Pizoli et al., 2002). The videos were captured to include the animals’ behavior before (baseline) and during each stimulation paradigm (ataxia, dystonia, and tremor). Each stimulation period lasted 2 min, during which the mouse was allowed to ambulate freely as well as in response to disturbance by the experimenter. The rating period excluded the first 10 s after stimulation was initiated to avoid including behavior in response to the sudden application of stimulation. Mice were given a score of 0 if no motor abnormalities were identified, 1 if there was abnormal motor behavior that was not obviously dystonic (such as severe tremor), 2 if there was a mild motor phenotype that included dystonic behaviors that may occur only in response to being disturbed by the experimenter or unprovoked (such as brief hyperextension of the limbs or extension of the neck and/or back), 3 if there was moderate severity or frequent unprovoked dystonic behaviors, 4 if there were severe unprovoked dystonic behaviors, and 5 if there were severe unprovoked dystonic behaviors that made goal-directed ambulation extremely difficult or impossible for an extended time. A detailed description of dystonic phenotypes in mice can be found in Brown et al. 2022 (23). Wilcoxon matched-pairs signed rank tests were performed with a post-hoc Holm-Sidak method for p-value adjustment in order to determine significance. Significance values were indicated as not significant (ns) if p > 0.05, * = p ≤ 0.05. ** = p ≤ 0.01.

Tremor analysis

The amplitude and frequency of tremor before (baseline) and during optogenetic stimulation was measured using a custom-made tremor monitor as previously described (20,36,78)The mice were implanted with optical fibers targeted to the interposed nucleus as described above. LED drivers were connected to the implant via optical patch cables and placed above the tremor monitor chamber. The tremor monitor chamber is a clear plexiglass box that is suspended by elastic cords that are connected to each corner of the box. An accelerometer is securely mounted to the bottom of the box to detect the tremor of the mouse within the box. The mice were placed in the chamber and were able to freely ambulate while in the box. They were given 2 min to acclimate to the novel tremor monitor environment before 2 min duration baseline recordings were made. The mice were then stimulated with the ataxia, dystonia, and tremor stimulation parameters for 2 min per stimulation period. At least 2 min were allowed to elapse between stimulation periods. Mice were encouraged to ambulate a similar amount during all recording periods. The output from the tremor monitor was amplified and lowpass filtered at 5 kHz (Brownlee Precision, Model 410) before being digitized by a CED board and analyzed using Spike2 scripts. Tremor monitor recordings were centered on zero and downsampled using the Spike2 interpolate function. A power spectrum analysis with a Hanning window was performed on each recording period. The peak power was calculated as the maximum power between 0 and 30 Hz, which is consistent with the range that we typically observe for physiological and pathological tremor in mice (20). A repeated-measures one-way ANOVA with a Tukey multiple comparison adjustment was performed to determine significance. Significance values were indicated as not significant (ns) if p > 0.05, * = p ≤ 0.05. ** = p ≤ 0.01.

Histology

After electrophysiology and behavior experimentation, mice were heavily anesthetized with avertin (2,2,2-tribromoethanol, Sigma-Aldrich, St. Louis, MO, USA; #T48402) and transcardially perfused with ice-cold phosphate buffered saline (PBS, 1x) followed by ice-cold 4% paraformaldehyde (PFA). The implants were removed, and then the brains were dissected out of the skull. The brains were post-fixed in PFA at 4°C for at least 24 hours. They were then cryoprotected via a sucrose gradient, starting at 15% sucrose in PBS followed by 30% sucrose in PBS. The brains were imbedded in Tissue-Tek O.C.T. Compound (Sakura, Torrance, CA, USA), frozen, and then sliced on a cryostat to produce 40 µm thickness sections. The tissue sections were washed in PBS, mounted onto slides, and allowed to dry on the slide for at least 2 hours. The sections were put in a xylene and ethanol series with ∼2 min per submersion in the following order: 100% xylene, 100% xylene, 100% ethanol, 100% ethanol, 90% ethanol, 70% ethanol. The sections were then placed in water for ∼2 min and then submerged in cresyl violet solution until the stain was sufficiently dark. They were then dehydrated by reversing the series with ∼30 s per submersion. Finally, Cytoseal XYL mounting media (Thermo Scientific, Waltham, MA, USA, #22-050-262) and a cover slip were placed on the slides. The slides were allowed to dry in a fume hood before imaging. Photomicrographs were obtained using a Leica DM4000 B LED microscope equipped with a Leica DMC 2900 camera and Leica Application Suite X (LAS X) software. Images were corrected for brightness and contrast using Adobe Photoshop (Adobe Systems, San Jose, CA, USA). Figure panels were made using Adobe Illustrator software.

Supporting information